---
title: "Statistical Phasing with External Beagle 5.x"
author: "HapBlockR Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Statistical Phasing with External Beagle 5.x}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(HapBlockR)
```

# Scientific distinction

An unphased dosage states how many alternate alleles an individual carries.
Phasing adds the chromosome assignment needed to identify alleles transmitted
together. Use validated phase for analyses of gametic haplotypes, cis
relationships or recombination; use dosage when allele count is the relevant
representation.

HapBlockR integrates Beagle 5.x as an external scientific dependency. The
user supplies the official JAR explicitly, preserving control of its version,
licence and execution.

# Configure Beagle

Download Beagle and the corresponding source from the official site, review
the GNU General Public License terms, and verify the downloaded file:

<https://faculty.washington.edu/browning/beagle/beagle.html>

Configure one explicit path:

```{r configure, eval=FALSE}
options(HapBlockR.beagle_jar = "/absolute/path/to/beagle.jar")

# Alternatively:
Sys.setenv(HAPBLOCKR_BEAGLE_JAR = "/absolute/path/to/beagle.jar")
```

The `beagle_jar` argument has highest priority, followed by the package
option, environment variable, and `beagle.jar` beside `out_prefix`.

Java 8 or later is required. Record `java -version` and the Beagle JAR
SHA-256 checksum with the analysis.

# Run and validate phasing

```{r run, eval=FALSE}
phased <- phase_with_beagle(
  input_vcf = "genotypes.vcf.gz",
  out_prefix = "results/genotypes_phased",
  nthreads = 2,
  seed = 42,
  ref_panel = "reference.vcf.gz",
  map_file = "genetic.map",
  min_genotype_concordance = 0.99,
  min_imputation_rate = 0.95,
  return_details = TRUE
)

phased$output_vcf
phased$quality_control
phased$provenance
```

Only one or two Beagle threads are accepted. HapBlockR verifies:

- the reported Beagle major version;
- input and output sample identity and order;
- chromosome, position, variant ID, REF, ALT, and variant order;
- dosage concordance at initially observed genotypes; and
- the imputation rate at initially missing genotypes.

Execution stops below user-defined thresholds. A structured provenance file
is written beside the phased VCF.

# Executable phased-VCF specification

This small deterministic example verifies the parser without requiring
Beagle during vignette construction.

```{r phased-fixture}
vcf <- tempfile(fileext = ".vcf")
writeLines(
  c(
    "##fileformat=VCFv4.2",
    paste(
      "#CHROM", "POS", "ID", "REF", "ALT", "QUAL",
      "FILTER", "INFO", "FORMAT", "P01", "P02",
      sep = "\t"
    ),
    paste("1", "100", "s1", "A", "G", ".", "PASS", ".",
          "GT", "0|1", "1|1", sep = "\t"),
    paste("1", "200", "s2", "C", "T", ".", "PASS", ".",
          "GT", "0|0", "1|0", sep = "\t")
  ),
  vcf
)

parsed <- read_phased_vcf(vcf)
parsed$hap1
parsed$hap2
parsed$dosage
stopifnot(
  identical(parsed$dosage, parsed$hap1 + parsed$hap2),
  identical(parsed$sample_ids, c("P01", "P02"))
)
```

# End-to-end pipeline

`run_ldx_pipeline(phase = TRUE)` accepts only VCF or VCF.gz input. It passes
the explicitly configured Beagle path, aligns phased output by immutable
variant and allele identity, and can cache phased matrices in a file-backed
`bigmemory` backend.

```{r pipeline, eval=FALSE}
result <- run_ldx_pipeline(
  geno_source = "genotypes.vcf.gz",
  out_dir = "results",
  out_blocks = "results/blocks.csv",
  out_diversity = "results/diversity.csv",
  out_hap_matrix = "results/haplotypes.csv",
  phase = TRUE,
  beagle_jar = "/absolute/path/to/beagle.jar",
  beagle_threads = 2,
  beagle_seed = 42,
  use_bigmemory = TRUE
)
```

# Interpretation and integration testing

Phased output remains model-dependent. Compare against a truth set when
possible and report switch error, genotype concordance, dosage accuracy,
imputation accuracy, and stratified performance by allele frequency.

The repository's required Beagle integration workflow downloads a pinned
Beagle 5.x release from the official source, records its checksum and Java
version, and prevents the phasing tests from silently skipping when that lane
is configured.

```{r session}
packageVersion("HapBlockR")
sessionInfo()
```
