---
title: "Large-Data Backends and Reproducible Performance Evidence"
author: "HapBlockR Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Large-Data Backends and Reproducible Performance Evidence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(HapBlockR)
set.seed(42)
```

# Memory model

HapBlockR provides dense, subset-oriented and file-backed access pathways.
The appropriate memory model depends on the input format and downstream
analysis.

| Pathway | Access model | Best use and memory implication |
| --- | --- | --- |
| GDS | Subset-oriented through SNPRelate | Downstream methods may still allocate dense matrices |
| PLINK BED | Operating-system memory mapping through BEDMatrix | Access pattern and downstream copies determine memory |
| `bigmemory` | File-backed matrix | Some algorithms materialise requested blocks |
| Text dosage or HapMap | Parsed text objects | Can require substantial in-memory representation |
| Ordinary VCF | Parser and optional GDS cache | Conversion, caching, and downstream objects have separate costs |

Peak memory depends on sample count, variant count, storage type, window size,
algorithm, missingness, threads, and copying by dependencies. Measure the
actual workflow on representative data.

# Executable correctness fixture

The small included data verify equality between one and two compiled threads.
This is a correctness test, not a performance claim.

```{r thread-equality}
data(ldx_geno)
x <- ldx_geno[seq_len(min(40L, nrow(ldx_geno))),
              seq_len(min(30L, ncol(ldx_geno))), drop = FALSE]

r_one <- compute_r2(x, n_threads = 1)
r_two <- compute_r2(x, n_threads = 2)

stopifnot(isTRUE(all.equal(r_one, r_two, tolerance = 1e-12)))
dim(r_one)
```

HapBlockR permits at most two compiled threads. Avoid nested parallel plans
that multiply process-level and compiled threads.

# File-backed workflows

Convert and validate data once, retain source and cache hashes, and preserve
sample and variant order.

```{r gds, eval=FALSE}
library(SNPRelate)

SNPRelate::snpgdsVCF2GDS(
  vcf.fn = "genotypes.vcf.gz",
  out.fn = "genotypes.gds",
  method = "biallelic.only"
)

gds <- read_geno("genotypes.gds")
```

```{r plink, eval=FALSE}
bed <- read_geno("genotypes.bed")
```

```{r bigmemory, eval=FALSE}
backed <- read_geno_bigmemory(
  "genotypes.vcf.gz",
  backingpath = "cache/genotypes"
)
```

HapBlockR cache manifests use SHA-256 source identity and parsing parameters.
A cache is reused only when that identity matches; otherwise it is refused or
explicitly rebuilt.

# Sparse and scalable LD options

Physical-distance restriction and community detection can reduce work:

```{r scalable-options, eval=FALSE}
blocks <- run_Big_LD_all_chr(
  geno_source = "genotypes.gds",
  CLQmode = "Leiden",
  max_bp_distance = 500000,
  subSegmSize = 1500,
  n_threads = 2
)
```

These options define the search space or segmentation method. Record their
biological rationale and compare them with a trusted scalar or smaller exact
analysis on a representative subset.

# Benchmark record

A publishable benchmark should retain:

```{r benchmark-schema}
benchmark_schema <- data.frame(
  field = c(
    "dataset_id", "input_sha256", "n_samples", "n_variants",
    "missing_rate", "method", "parameters", "threads", "R_version",
    "HapBlockR_version", "compiler", "hardware", "wall_seconds",
    "cpu_seconds", "peak_resident_mb", "replicate", "correctness_passed"
  ),
  required = TRUE
)
benchmark_schema
```

Report medians and variability from repeated runs, and distinguish measured
results from extrapolations. Promote a performance result after numerical
equality, sample identity and allele identity have passed.

# Scheduled integration

Large public or simulated fixtures belong in scheduled CI, with:

- fixed data checksums and package versions;
- hardware and compiler records;
- one- versus two-thread equality;
- a trusted scalar comparison;
- peak resident memory;
- performance non-regression thresholds; and
- retained benchmark artefacts.

The examples above remain non-executing only where they require external
files or optional backends; the vignette's correctness and schema examples
are executed on every build.

```{r session}
packageVersion("HapBlockR")
sessionInfo()
```
