---
title: "From Local GEBV to a Crossing Decision: Haplotype-Based Parent Selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From Local GEBV to a Crossing Decision: Haplotype-Based Parent Selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  message    = FALSE,
  warning    = FALSE
)
library(HapBlockR)
data(ldx_geno,     package = "HapBlockR")
data(ldx_snp_info, package = "HapBlockR")
data(ldx_blocks,   package = "HapBlockR")
data(ldx_blues,    package = "HapBlockR")
```

## 1. The question this vignette answers

The other vignettes in this package are mostly about *finding structure* in
marker data: detecting linkage disequilibrium (LD) blocks, extracting
haplotypes, measuring diversity. This one is about *using that structure to
make a breeding decision*.

The starting point here is deliberately downstream of where most genomic
prediction guides stop: you already have a breeding value or selection index
for every candidate parent — from a mixed model, Genomic Best Linear
Unbiased Prediction (GBLUP), ASReml-R, or any other method of your choosing.
HapBlockR does not replace that step. What it adds is everything between
"here is a ranked list of candidates" and "here is the crossing block for
next season," specifically:

1. **Where** in the genome is that breeding value actually coming from
   (local GEBV, Genomic Estimated Breeding Value, per haplotype block, not
   just one genome-wide number)?
2. **Which set** of parents, taken together, covers the most favourable
   haplotype blocks — and how is that different from simply taking the top
   N individuals by whole-genome value?
3. **Is that set balanced** across your programme's families, or does one
   family dominate the crossing block?
4. **Does it actually pay off** — if you found a recurrent-selection programme
   on this set instead of the naive top-N list, is realised genetic gain over
   generations actually higher?
5. **Which specific crosses**, contribution levels, and matings turn that
   parent set into an actual crossing block — with population-wide
   inbreeding under an explicit cap, not just a diversity check run
   afterward?

Each section below answers one of these questions, in order, ending with an
actual list of parents (Sections 3-7) and then, building on that shortlist,
a full cross-and-mating decision with its own diversity and validation
checks (Sections 8-12) — the evidence for why each choice was made over
the obvious alternative, at every step.

---

## 2. Illustrative scenario

This vignette reuses the same example data as the introductory vignette --
`ldx_geno` (120 individuals x 230 SNPs across 3 chromosomes), the
pre-detected `ldx_blocks` (9 LD blocks plus inter-block singleton SNPs), and
`ldx_blues` (pre-adjusted BLUEs for two simulated traits, `YLD` and `RES`).
See the *Introduction* vignette for how the blocks themselves were detected;
this vignette starts one step later, from a genotyped, block-mapped panel
with a breeding value already in hand.

To illustrate the family-balance check in Section 6, this vignette also adds
a synthetic `Family` grouping — 12 illustrative families of 10 lines each.
**This grouping is invented purely for this example.** Your own programme would
use its actual pedigree/family records here instead.

```{r add-family}
set.seed(1)
family_id <- sample(rep(paste0("Fam", sprintf("%02d", 1:12)), each = 10))
names(family_id) <- rownames(ldx_geno)
table(family_id)
```

`YLD` stands in for a pre-computed selection index (higher = better):

```{r index}
index_df <- ldx_blues[, c("id", "YLD")]
names(index_df)[2] <- "SelectionIndex"
head(index_df)
```

---

## 3. Step 1: Where is the value coming from?

A single genome-wide breeding value tells you *how good* a candidate is, not
*where* that value comes from. `run_haplotype_prediction()` re-expresses the
same predicted value as a sum of per-block contributions — local GEBV — so
you can see which specific haplotype blocks are actually driving
performance, following the haplotype-stacking methodology of Tong et al.
(2025, *Theor Appl Genet* 138:267). This is the first step toward a
genomics-informed crossing decision rather than a black-box ranking.

```{r run-prediction}
pred <- run_haplotype_prediction(
  geno_matrix = ldx_geno,
  snp_info    = ldx_snp_info,
  blocks      = ldx_blocks,
  blues       = index_df,
  id_col      = "id",
  blue_col    = "SelectionIndex",
  marker_effect_method   = "gblup",
  complete_decomposition = TRUE,
  verbose     = FALSE
)

pred$n_blocks                                    # includes singleton pseudo-blocks
sum(pred$block_importance$important)              # blocks clearing the 0.9 cutoff
head(pred$block_importance[
  order(-pred$block_importance$var_scaled),
  c("block_id", "CHR", "n_snps", "var_scaled", "important")
])
```

In most panels, a handful of blocks account for most of the variance in
local GEBV — these are the "vital few" a breeder actually needs to track and
stack. The rest contribute comparatively little on their own.

---

## 4. Step 2: Separate the vital few from the trivial many

`plot_block_funnel()` makes that split visible: block-level local GEBV on the
x-axis, scaled variance on the y-axis, with the important-block threshold
highlighted.

```{r funnel-plot, fig.alt="Funnel plot of per-block local GEBV against scaled variance, with important blocks highlighted above the 0.9 threshold line"}
plot_block_funnel(
  local_gebv          = pred$local_gebv,
  block_importance    = pred$block_importance,
  highlight_threshold = 0.9,
  max_blocks          = 2000L
)
```

`select_top_blocks()` turns that visual judgement into the smallest top-ranked
set of blocks explaining a target share of the variance — the actual target
list for the parent-selection step below, rather than the whole genome:

```{r top-blocks}
top_blocks <- select_top_blocks(
  block_importance  = pred$block_importance,
  n                 = NULL,
  perc_total        = NULL,
  perc_of_total_var = 0.90,
  var_col           = "var_scaled"
)

nrow(top_blocks)
top_blocks[, c("block_id", "CHR", "n_snps", "var_scaled", "cum_var_share")]
```

---

## 5. Step 3: Which parents actually cover these blocks?

There are two competing answers to "who are our best 20 parents," and this
package deliberately implements both so they can be checked against each
other rather than trusting either blindly.

### 5.1 Truncation selection — the baseline

`truncation_selection(score, n_founders)` does exactly what its name says
and nothing more: rank every candidate by a single numeric score (whole-genome
breeding value, a stacking index, anything you supply), take the top
`n_founders`. There is no block-awareness, no notion of complementarity
between chosen parents, and no diversity or relatedness management --
two of the top-20 individuals by score could be full sibs, or could all carry
the exact same favourable haplotypes and none of the others. Its value is
precisely that simplicity: it is the default assumption every smarter method
needs to beat, and if a smarter method can't beat it, that is itself
important information.

**When to reach for it.** Use `truncation_selection()` whenever you have one
already-validated selection index and no open complementarity question --
early-stage or small programmes without the infrastructure for a full
haplotype-block workflow; time-pressured decisions (a release deadline, a
seed increase) where a fast, transparent, easy-to-explain shortlist matters
more than incremental optimisation; and, always, as the mandatory baseline
you run *every time* alongside a smarter method (Section 5.2 onward), so you
have evidence for whether the extra machinery actually earned its keep on
this panel.

**When to look further instead.** If your elite candidates cluster tightly
on a handful of pedigrees, a top-N-by-value list is likely to also be a
top-N-by-relatedness list — Section 6's family-balance check will catch
this after the fact, but if you already expect it, go straight to Section
5.2. If you've already identified specific favourable haplotype blocks
worth stacking deliberately (Sections 3-4), truncation selection has no way
to act on that information at all.

**In practice.** One function call: `score` is whatever whole-genome
value you already trust (GEBV, a selection index, a stacking index from
`score_favorable_haplotypes()`), `n_founders` is your programme's crossing-
block capacity. No relatedness matrix, no block data, no tuning. The output
is a flat character vector of IDs (`$selected`) — usable immediately, or
as the comparison arm for every other section below.

### 5.2 GA parent selection — optimising for joint block coverage

`select_parents_ga()` answers a narrower, different question: not "who
individually scores highest," but "which *set* of `n_founders` individuals,
taken together, covers the most favourable value across the target blocks
from Step 2." Concretely, its fitness function is
$\sum_j w_j \cdot \mathrm{best}_j(S)$ — for every target block $j$,
take the best value the chosen set $S$ can achieve at that block (see the
`strategy` table below for what "achieve" means), weight it by $w_j$
(`block_weights`), and sum across all target blocks. A binary genetic
algorithm (`GA::ga(type = "binary")`) searches over which individuals to
include. This is what turns "here is a ranked list of good candidates" into
"here is an actual founder set engineered to jointly carry your target
haplotypes" — something no amount of eyeballing a truncation-selection list
can guarantee, since the top-N by whole-genome value can easily be
concentrated on the *same* favourable blocks while leaving others uncovered.

`select_parents_ga()`'s `strategy` argument encodes how a block's value is
"delivered" by the chosen set, mapped onto real crossing schemes:

| `strategy` | Breeding meaning |
|---|---|
| `"no_selfing"` (default), `"OHS"` | Two distinct parents must jointly carry the block (standard biparental cross) |
| `"selfing"`, `"Haploid_OHS"` | One parent alone is enough (selfing, or a doubled-haploid-style single-parent gamete donation) |
| `"OPV"` | One parent carrying the block is enough, population-value framing (no specific pairing required) |

**When to reach for it.** Use `select_parents_ga()` once Step 1-2 has given
you a specific list of target haplotype blocks worth stacking simultaneously
-- this is the tool for "which set of individuals, taken together, covers
what I actually want covered." It earns its cost particularly when: elite
candidates in your programme carry overlapping favourable regions (truncation
selection would double up on the same blocks and leave others uncovered);
you are assembling a multi-parent founder population (MAGIC, NAM, a new
recurrent-selection base) where joint complementarity matters more than any
single individual's rank; or you need a specific crossing-scheme assumption
(no_selfing, OHS, OPV, Haploid_OHS) enforced *during* the search rather than
checked afterward.

**When to skip it (for now).** You need `local_gebv` from
`run_haplotype_prediction()` first — there is no version of this tool that
works from a single whole-genome score alone (that's Section 5.1). For very
large candidate pools (thousands of individuals) without a `top_candidates`
pre-filter, GA search time becomes a real cost worth budgeting for. And if
Section 5.1's truncation list already looks complementary (Section 5.2's
own overlap check, below, tells you this directly), the added workflow step
may not change the outcome enough to justify it every cycle.

**In practice.** You need `value_matrix` (candidates x target blocks, sliced
straight from `local_gebv`), `n_founders`, `strategy`, and `block_weights`
(typically each block's `var_scaled`). Always pair it with a `merit_score`/
`min_sel_value` floor — without one, the search can select a genuinely poor
overall performer purely for uniquely covering one block, since the fitness
function itself has no whole-genome-merit term. Run with `n_reps >= 5` for
a real analysis (Section 5.3 explains why) and always compare against
`truncation_selection()`'s overlap, not GA output in isolation.

Both functions also take a `min_sel_value`/`min_sel_mode` merit floor,
applied to the candidate pool *before* ranking or searching. Without it,
`select_parents_ga()` can select a genuinely poor overall performer purely
because they uniquely cover one target block — its fitness function has no
term for whole-genome merit at all (see `merit_score` below, which supplies
the value the floor is evaluated against). `min_sel_mode = "percentile"`
keeps the floor self-scaling: `0.5` keeps the top half of candidates by
`merit_score`/`score`, regardless of what units or range that score is on.

```{r ga-vs-ts}
value_matrix <- pred$local_gebv[, top_blocks$block_id, drop = FALSE]

ga_sel <- select_parents_ga(
  value_matrix   = value_matrix,
  n_founders     = 20L,
  strategy       = "no_selfing",
  block_weights  = top_blocks$var_scaled,
  top_candidates = NULL,
  popSize        = 100L,
  maxiter        = 200L,
  run            = 50L,
  pmutation      = 0.1,
  pcrossover     = 0.8,
  penalty_weight = NULL,
  seed           = 1L,
  verbose        = FALSE,
  merit_score    = pred$gebv,      # whole-genome value the floor is judged on
  min_sel_value  = 0.5,            # keep the top 50% by merit_score
  min_sel_mode   = "percentile",
  n_reps         = 2L              # reduced for vignette build speed; the
                                    # package default is 5 — see below
)

ts_sel <- truncation_selection(score = pred$gebv, n_founders = 20L,
                               min_sel_value = 0.5, min_sel_mode = "percentile")

length(intersect(ga_sel$selected, ts_sel$selected))   # parents both methods agree on
sort(ga_sel$selected)
```

The overlap count is informative on its own: a high overlap means truncation
selection was already close to the GA's answer for this panel; a low overlap
means block coverage and whole-genome ranking disagree, which is worth
investigating further — it usually means the top-ranked individuals by
whole-genome GEBV are concentrated on the *same* favourable blocks, leaving
other important blocks uncovered by a plain top-N list.

### 5.3 Is the GA's answer a stable optimum?

A single GA run says nothing about whether its answer is a robust optimum or
one of several near-equally-good solutions a stochastic search happened to
land on. `select_parents_ga()` addresses this directly: with `n_reps`
independent replicates (package default `5`; `2` above only to keep this
vignette fast to build), `$stability` reports how consistent the answer was:

```{r ga-stability}
ga_sel$converged                          # did the winning replicate plateau?
ga_sel$stability$fitness_range            # best fitness, across replicates
sort(ga_sel$stability$selection_freq, decreasing = TRUE)[1:10]
```

`selection_freq` is the fraction of replicates that selected each individual
-- candidates at `1.0` are robustly supported regardless of the GA's random
starting point; anyone selected in only one replicate out of several is
borderline and worth a second look before committing to them. Increase
`n_reps` for a real analysis; `2` here is a vignette-build-speed compromise,
not a recommended value.

### 5.4 What neither method does

Both methods answer a real question, but neither answers every question a
finalised crossing plan needs. Specifically, **neither**:

- **manages coancestry or inbreeding risk** in the chosen set — `select_top_blocks()`/
  `select_parents_ga()`/`truncation_selection()` have no relatedness penalty.
  The family-balance check in Step 4 below is a diagnostic you run
  *afterward*, not a constraint the search itself respects;
- **assigns differential contributions** — both return a flat set of
  `n_founders` individuals with no notion that some parents should
  contribute more crosses than others;
- **decides who mates whom** — neither produces an actual cross list, only
  a candidate/founder set.

Dedicated optimal-contribution-selection (OCS) and mate-allocation tools
(e.g. AlphaMate, Gorjanc lab) are built specifically to solve those three
problems jointly — typically maximising expected genetic gain subject to an
explicit group-coancestry constraint, then allocating differential
contributions and actual matings across the constrained set. A practical
combined workflow is to use `select_top_blocks()`/`select_parents_ga()` to
identify *which candidates are worth considering* (i.e. those covering your
programme's important haplotype blocks), then hand that shortlist to a
dedicated OCS tool for the coancestry-managed contribution and mating
decision, rather than treating HapBlockR's parent selection as a final
mating plan on its own.

### 5.5 How the shortlist stage and the downstream stage actually relate

It's worth being explicit about the shape of this workflow, since it's easy
to misread it as a straight pipeline (truncation selection -> GA selection
-> UC -> OCS) when it is not one. Two genuinely separate layers are at
work:

- **The shortlist layer** — `truncation_selection()`, `select_parents_ga()`,
  and `select_parents_by_family()` — are *alternatives to each other*, not
  a sequence. Each answers "who are my candidate parents" by different
  logic (single-score ranking, block-coverage search, or group quotas). The
  normal workflow is to run more than one and compare (Section 5.2's
  overlap check, Section 6's family-balance check), not to feed one's
  output into another.
- **The downstream layer** — `usefulness_criterion()` (Section 8) and
  `select_parents_ocs()` (Section 9) — sits one stage further on, and
  neither builds its own candidate pool. Both are written to consume
  *whichever* flat shortlist you settled on, plus that shortlist's merit
  values (and, for `select_parents_ocs()`, `G`). This vignette's examples
  below happen to hand them `ga_sel$selected` purely for narrative
  continuity — that is not a requirement of either function.
  `ts_sel$selected`, a `select_parents_by_family()` result, or a
  hand-curated list all work identically, since neither downstream function
  inspects or cares which upstream tool (or process) produced its input.

`usefulness_criterion()` and `select_parents_ocs()` are themselves siblings
at that downstream layer, not a further sequence — cross ranking (Section
8) scores candidate pairs independently, with no notion of an overall
population-wide plan; `select_parents_ocs()` (Section 9) solves a
genuinely different problem (contributions and matings under an explicit
relatedness cap) directly from merit and `G`, and does not take
`usefulness_criterion()`'s ranked-cross table as an input at all. You can
use Section 8's ranking to eyeball which pairings look promising before
committing to a Section 9 run, but that is a manual judgement call, not a
data dependency between the two functions.

**One important qualification to "downstream, consumes whatever you hand
it."** That description is accurate for *where candidates come from* --
`select_parents_ocs()` can never select a parent that wasn't in the
`merit`/`G` list you supplied, regardless of engine. It is not accurate
to read as "the mating plan necessarily uses every candidate in your
shortlist unchanged." None of the three engines is obliged to give every
supplied candidate a nonzero contribution, and `n_parents_max` makes this
an explicit, deliberate step rather than an incidental one — and the three
engines implement it very differently: AlphaMate enforces it *natively,
inside* its evolutionary algorithm, jointly deciding which subset of your
candidates contributes at all, how much, and who mates with whom, in one
optimisation; `"optisel"` instead solves the continuous contribution optimum
over your *entire* supplied list first, then, only as a post-hoc,
non-re-optimised patch, zeros out every contribution below the
`n_parents_max`-th largest (see *Engine differences* in Section 9); and
`"simplemating"` does not support `n_parents_max` at all (ignored, with a
message), though its own greedy cross selection can still leave some
supplied candidates out of the final plan incidentally. So: the shortlist
you hand `select_parents_ocs()` is a hard ceiling on who can appear in the
mating plan, not a guarantee that everyone in it will.

### 5.7 Decision guide

| Question you're asking | Use |
|---|---|
| "What's the simplest defensible shortlist?" | `truncation_selection()` |
| "Are my best-by-value individuals actually covering my target haplotype blocks, or double-counting the same ones?" | `select_parents_ga()`, compared against `truncation_selection()` |
| "Is one method clearly better here, or do they agree?" | Run both, check `intersect()` (Section 5.2) — low overlap means block coverage and whole-genome ranking disagree and both are worth a look |
| "Which specific parents should I be most confident about?" | Parents both methods select (a high-overlap subset), and see Step 4 for whether that subset is family-balanced |
| "I'm worried block-coverage search is selecting on noisy per-block estimates" | `select_parents_ga(merit_weight = ...)`, or the easier `merit_priority = ...` (0-100, auto-calibrated — see `suggest_merit_weight()`) — Section 5.8 |
| "I want a relatedness cap on the GA's founder set, without guessing a raw `coancestry_weight`" | `select_parents_ga(target_degree = ...)` (0-90, same convention as `select_parents_ocs()`) — Section 5.8 |
| "My programme's shortlist is naturally 'best few families, best few lines each'" | `select_parents_by_family()` — Section 5.8 |
| "I need contribution numbers and an actual mating list, with inbreeding under control" | Neither — shortlist candidates here, then use a dedicated OCS/mate-allocation tool (Section 5.3) |

### 5.8 Two further shortlist strategies: merit-weighted GA, and family quotas

Sections 5.1-5.2 leave two gaps a real programme's shortlist decision often
runs into. First, `select_parents_ga()`'s block-coverage search (Section
5.2) has no notion of whole-genome merit unless `min_sel_value` pre-filters
the candidate pool — and even then, a candidate that barely clears the
floor is treated identically to one that clears it by a wide margin. Since
per-block local GEBV are themselves statistical estimates, not ground
truth, a coverage-only search can end up leaning on a candidate whose
apparent block coverage is partly a noisy artefact. Setting `merit_weight`
adds a second, additive term to the same fitness function used in Section
5.2 — whole-genome merit now competes with block coverage throughout the
search, not just at the entry gate:

```{r merit-weight-ga}
ga_hybrid <- select_parents_ga(
  value_matrix   = value_matrix,
  n_founders     = 20L,
  strategy       = "no_selfing",
  block_weights  = top_blocks$var_scaled,
  merit_score    = pred$gebv,
  merit_weight   = 0.5,            # tune against $fitness's typical scale
  min_sel_value  = 0.5,
  min_sel_mode   = "percentile",
  seed           = 1L,
  verbose        = FALSE,
  n_reps         = 2L              # reduced for vignette build speed
)

ga_hybrid$mean_merit                       # realised mean merit_score of the chosen set
mean(pred$gebv[ga_sel$selected])           # vs. Section 5.2's coverage-only run
```

Compare `ga_hybrid$mean_merit` against the coverage-only `ga_sel` run from
Section 5.2 (and against `truncation_selection()`'s own mean merit) to see
the term's effect directly — there is no universal correct `merit_weight`,
the same tuning approach as `coancestry_weight` (Section 5.2) applies.

Guessing `merit_weight = 0.5` above and inspecting the result afterward
works, but there is an easier way: `merit_priority` states how much you
care about merit vs. coverage as a plain 0-100 percentage, and HapBlockR
calibrates the matching `merit_weight` from your own data automatically.
`suggest_merit_weight()` runs the exact same calibration standalone, so
you can see the numbers first:

```{r merit-priority}
cal <- suggest_merit_weight(
  value_matrix   = value_matrix,
  merit_score    = pred$gebv,
  n_founders     = 20L,
  strategy       = "no_selfing",
  block_weights  = top_blocks$var_scaled,
  merit_priority = 50                    # "half as much weight as coverage"
)
cal$merit_span; cal$coverage_span; cal$suggested_merit_weight

ga_dial <- select_parents_ga(
  value_matrix   = value_matrix,
  n_founders     = 20L,
  strategy       = "no_selfing",
  block_weights  = top_blocks$var_scaled,
  merit_score    = pred$gebv,
  merit_priority = 50,              # calibrated internally — no merit_weight guess
  min_sel_value  = 0.5,
  min_sel_mode   = "percentile",
  seed           = 1L,
  verbose        = FALSE,
  n_reps         = 2L
)
ga_dial$merit_weight               # the value merit_priority = 50 resolved to
```

What `suggest_merit_weight()` actually calibrates: for a group of size
`n_founders`, the realistic best-vs-worst achievable *spread* of each term
-- coverage's ceiling built via a real, feasible greedy-selected group (not
an unreachable per-block best-value sum), both floors trimmed against a
single outlier candidate distorting the estimate. `merit_priority = 100`
would scale merit's spread to match coverage's; `50` splits the difference.
This is one reasonable, clearly-stated definition of "comparable" — not
the only one — so treat the suggestion as an informed starting point, not
a uniquely correct answer; `merit_weight` remains available whenever you
want to fine-tune by hand instead. `merit_weight` and `merit_priority` are
mutually exclusive on a single `select_parents_ga()` call — pick one.

`coancestry_weight` (introduced in Section 5.8/10) has the same scale
problem, and `select_parents_ga()` has an easier alternative for it too:
`target_degree`, on the SAME `[0, 90]` scale `select_parents_ocs()` already
uses in Section 9 (`0` = max gain, `90` = max diversity) — one consistent
dial across the whole package, rather than a differently-shaped one per
function. The mechanism is genuinely different from `select_parents_ocs()`'s
own `target_degree`, though: that function solves the real OCS
contribution-optimisation problem at each frontier extreme; this one
converts your `[0, 90]` choice into a relatedness *ceiling* — interpolated
between a greedy high-coverage reference group and
`select_core_collection(strategy = "maximin")`'s diversity-maximising one
-- then enforces it as a penalty inside the same block-coverage GA:

```{r target-degree}
ga_degree <- select_parents_ga(
  value_matrix   = value_matrix,
  n_founders     = 20L,
  strategy       = "no_selfing",
  block_weights  = top_blocks$var_scaled,
  G              = pred$G,
  target_degree  = 30,              # same convention as select_parents_ocs()
  seed           = 1L,
  verbose        = FALSE,
  n_reps         = 2L
)
ga_degree$relatedness_ceiling      # the ceiling target_degree = 30 resolved to
ga_degree$mean_relationship        # the chosen set's actual realised relationship
```

`coancestry_weight` and `target_degree` are mutually exclusive on a single
`select_parents_ga()` call, exactly like `merit_weight`/`merit_priority` --
pick one.

Second, many programmes' real shortlist decision already has a different
shape than either "top-n by value" or "GA-searched coverage set": pick the
best few *groups*, then the best few *lines* from each. A "group" is either
your own pedigree/family labels (the default, used below) or a
data-derived genetic cluster built directly from a relationship matrix
(`group_by = "genetic_cluster"`, for programmes where pedigree labels aren't
a fully faithful proxy for actual relatedness — see
`?select_parents_by_family`'s "Grouping: pedigree family vs. genetic
cluster" section). `select_parents_by_family()` implements the
best-groups-then-best-lines shape, using the same synthetic `Family`
grouping introduced in Section 2 (substitute your own pedigree data here).
Group ranking is shrinkage-corrected by default (a group's raw top-`rank_k`
mean is unreliable when the group is small — empirical-Bayes/BLUP-style
family evaluation shrinks it toward the across-group mean by a reliability
weight estimated from every eligible member's own score; pass
`family_rank_method = "topk_mean"` to restore the historical, unshrunk
ranking). Its optional `ensure_haplotype_diversity = TRUE` mode takes the
allele-level detail from `extract_haplotypes()` directly — whatever object
your pipeline already produced — and reshapes it internally, so no manual
matrix construction is needed on your end:

```{r family-quota-selection}
haps <- extract_haplotypes(ldx_geno, ldx_snp_info, ldx_blocks, min_snps = 3L)

fam_sel <- select_parents_by_family(
  score         = pred$gebv,
  family        = family_id,
  n_families    = 6L,
  n_per_family  = 3L,
  ensure_haplotype_diversity = TRUE,
  value_matrix  = value_matrix,
  haplotypes    = haps,           # auto-derives allele identity per block
  diversity_method = "dominant_block"  # allele-level check needs `haplotypes`;
                                        # the new default, "coverage_gain",
                                        # ignores `haplotypes` — see below
)

fam_sel$family_ranking[, c("family", "topk_mean", "shrinkage_weight", "rank_score", "n_members", "rank", "selected")]
fam_sel$by_family[, c("family", "individual", "score", "dominant_block", "collision")]
```

`ensure_haplotype_diversity = TRUE` adjusts within-family picks (greedily,
family by family in rank order) to prefer lines that add something new
relative to every pick already claimed by a higher-ranked family, so that
crossing a representative of one chosen family with a representative of
another is more likely to combine two different favourable haplotypes
rather than duplicate one; `by_family$collision` flags any pick where no
alternative existed and merit rank (not the diversity preference) decided
the outcome. Two `diversity_method` choices implement "adds something new"
differently: `"dominant_block"` (used above, for the allele-level check
against `haplotypes`) looks only at each line's single highest-value
target block; the new default, `"coverage_gain"`, instead reuses
`select_parents_ga()`'s own block-coverage machinery to ask whether a
candidate's addition genuinely increases total value-weighted coverage
across *every* target block given everyone already claimed — a more
rigorous check, at the cost of ignoring `haplotypes`' allele-level detail.
This is a lighter-weight, group-quota-scoped alternative to
`select_parents_ga()`'s joint block-coverage search. A relationship matrix
`G` also unlocks an optional relatedness ceiling within each group
(`within_group_target_degree`, the same 0-90 convention as
`select_parents_ga()`'s `target_degree`) and, together with `n_clusters`,
genetic-cluster grouping itself — see `?select_parents_by_family`'s
"When to reach for this instead of truncation_selection() or
select_parents_ga()" section for the full decision guide between all three
founder-shortlist strategies.

---

## 6. Step 4: Is either selection a diversity or family risk?

`plot_parent_selection_pca()` projects every candidate onto a genomic PCA,
coloured by selection group:

```{r pca-plot, fig.alt="PCA scatterplot of all 120 candidates coloured by whether they were GA-selected, truncation-selected, both, or neither"}
plot_parent_selection_pca(
  G           = pred$G,
  ga_selected = ga_sel$selected,
  ts_selected = ts_sel$selected
)
```

`select_parents_ga()` has **no built-in family or pedigree constraint** — it
optimises purely on haplotype-block coverage. The family-balance table below
is the manual check a breeder should run before committing to a crossing
block, using the synthetic `Family` grouping from Section 2 (substitute your
own pedigree data here):

```{r family-balance}
family_summary <- function(selected_ids, label) {
  tab <- sort(table(family_id[selected_ids]), decreasing = TRUE)
  cat(label, "-- top family:", names(tab)[1], "with",
      round(100 * tab[1] / length(selected_ids)), "% of selected parents\n")
  tab
}

family_summary(ga_sel$selected, "GA selection")
family_summary(ts_sel$selected, "Truncation selection")
```

If one family dominates either set, that is a diversity/inbreeding-risk
signal to catch here — before the crosses are made — not after.

---

## 7. Step 5: Does the GA selection actually deliver more genetic gain?

The previous two sections compare the two founder sets *structurally*
(coverage, overlap, diversity). This section is the "prove it" step: found a
recurrent-selection programme on each set and compare realised whole-genome
GEBV over generations.

`ga_vs_ts_simulation()` requires **phased haplotypes** (`hap1`/`hap2`) --
not shipped as example data in this package, since the shipped `ldx_geno` is
unphased dosage. See the *Statistical Phasing* vignette for
`run_ldx_pipeline(phase = TRUE)` or `read_phased_vcf()`. It also wraps the
[genomicSimulation](https://github.com/vllrs/genomicSimulation) package,
which is not on CRAN — see the Installation section of the README for the
recommended install order. The code below is shown for reference and is not
executed in this vignette, since it needs both of those prerequisites:

```{r ga-vs-ts-sim, eval=FALSE}
# phased$hap1 / phased$hap2: SNPs x individuals, from read_phased_vcf()
# or run_ldx_pipeline(phase = TRUE)
phased <- read_phased_vcf("mydata_phased.vcf.gz")

sim <- ga_vs_ts_simulation(
  hap1                 = phased$hap1,
  hap2                 = phased$hap2,
  snp_info             = phased$snp_info,
  snp_effects          = pred$snp_effects,
  ga_selected          = ga_sel$selected,
  ts_selected          = ts_sel$selected,
  n_generations        = 10L,
  pop_size             = 100L,
  recomb_rate          = 1e-8,
  selection_intensity  = 0.2,
  seed                 = 1L,
  verbose              = FALSE
)

plot_ga_vs_ts_simulation(sim, show_max = TRUE)

# Per-generation mean/max GEBV for each scheme:
sim$summary[sim$summary$scheme == "GA", ]
sim$summary[sim$summary$scheme == "TS", ]
```

`sim$summary` gives `mean_gebv` and `max_gebv` per generation per scheme
(generation 0 = the founders themselves). The pattern to look for is whether
the GA-founded population's `mean_gebv` trajectory pulls ahead of the
TS-founded one as generations accumulate — this is expected to be most
pronounced exactly when Section 5's founder-set overlap was low, since that
is precisely when block coverage and whole-genome ranking disagreed about
which parents to advance. When the two founder sets overlap heavily, expect
the two trajectories to track closely as well, since the underlying genetic
potential loaded into each scheme was already nearly the same.

**Beyond founder set: comparing *mating strategies*, not just founders.**
The comparison above holds the mating rule fixed (random-mate, then keep the
top `selection_intensity` fraction by GEBV — "truncation" selection within
each generation) and varies only which founders start the programme. A
separate, complementary question is whether a smarter *mating* rule beats
truncation outright, holding founders fixed. The `schemes` argument
generalises `ga_vs_ts_simulation()` to any number of named schemes, each
with its own `mating_scheme`: `"truncation"` (the default above), `"ocs"`
(each generation, `select_parents_ocs()` picks an optimum-contribution
mating plan instead of random-mating), or `"uc"` (each generation,
`usefulness_criterion()` ranks every candidate pair and only the top-ranked
ones are crossed; requires `blocks`, the LD-block table from Section 1):

```{r ga-vs-ts-sim-schemes, eval=FALSE}
sim4 <- ga_vs_ts_simulation(
  hap1          = phased$hap1,
  hap2          = phased$hap2,
  snp_info      = phased$snp_info,
  snp_effects   = pred$snp_effects,
  schemes = list(
    GA  = list(founders = ga_sel$selected, mating_scheme = "truncation"),
    TS  = list(founders = ts_sel$selected, mating_scheme = "truncation"),
    OCS = list(founders = ts_sel$selected, mating_scheme = "ocs",
              n_crosses = 20L),
    UC  = list(founders = ts_sel$selected, mating_scheme = "uc",
              selected_proportion = 0.1)
  ),
  blocks        = blocks,       # only needed because the "uc" scheme is here
  n_generations = 10L,
  seed          = 1L,
  verbose       = FALSE
)

plot_ga_vs_ts_simulation(sim4)
```

This answers a genuinely different question from the GA-vs-TS founder
comparison: given the *same* starting parents, does managing the
diversity-vs-gain tradeoff every generation (OCS) or targeting the
best-predicted crosses every generation (UC) outperform simply truncating
by GEBV? `ga_selected`/`ts_selected` remain available exactly as shown
above for the simpler 2-scheme case — `schemes` is purely additive, not a
breaking change.

---

## 8. Step 6: Which crosses among these parents are worth making?

Sections 3-7 answer "who are the parents." They stop short of "who should be
crossed with whom" — `select_parents_ga()`/`truncation_selection()`/
`select_parents_by_family()` each return a flat set with no notion of
pairing. `usefulness_criterion()` (the Usefulness Criterion / UC, Schnell &
Utz 1975; Bernardo 2003; Zhong & Jannink 2007) picks up from there and
ranks candidate **crosses**, not candidate parents, by the predicted
quality of the best progeny a cross could plausibly produce — it takes
whichever of those three shortlists (or any parent list at all) you hand
it as `parent_ids`, with no preference for one origin over another (see
Section 5.5):
$$UC = \mu_{\text{mid-parent}} + i \cdot \sqrt{\sigma^2_{\text{cross}}}$$
-- mid-parent value plus a selection-intensity-scaled term for the cross's
predicted genetic variance. A high-UC cross does not need high-merit parents
on both sides; a cross between a strong parent and a complementary
weaker-but-different one can out-rank two mediocre-but-similar strong
parents, because it is predicted to segregate a better best-progeny tail.

Four variance-prediction modes trade off data requirements against rigor
(see `?usefulness_criterion` for full detail):

| `variance_model` | Requires | What it captures |
|---|---|---|
| `"block_independent"` (shown below) | `local_gebv` (already in hand from Section 3) | Single-locus segregation variance per target block, summed across blocks — works on unphased data, coarser for partially heterozygous parents |
| `"phased"` | Phased haplotypes (`extract_haplotypes()` on phased input) | Exact 4-gamete enumeration per block — exact within a block, still treats blocks as independent |
| `"linked"` | Same phased haplotypes as `"phased"`, plus a `genetic_map` (no extra package) | Monte Carlo progeny simulation (Haldane-mapped block-to-block recombination) — closes `"phased"`'s independent-blocks gap natively, without `SimpleMating`; still single-cross F1-style variance, not a multi-generation population variance |
| `"simplemating"` | 0/2-coded genotypes (DH/RIL), a genetic map or LD matrix | Genuine multi-locus Mendelian-sampling covariance via `SimpleMating::getUsefA()` (Peixoto et al. 2024) — the most rigorous of the four, since it also models multi-generation RIL/DH population variance, not just a single cross |

**When to reach for it.** Once you have a founder/parent shortlist (from
Sections 5.1/5.2, or your own list built any other way), `usefulness_criterion()`
is the tool for the next question down: not "who do I keep," but "which
*pairs* do I actually cross." It is particularly worth running when your
programme only has capacity to make a limited number of biparental crosses per
cycle and wants to prioritise the ones most likely to segregate a genuinely
better best-progeny tail, not just the ones between the two highest-ranked
individuals; or when you suspect a complementary-but-moderate pairing might
out-perform two similar elite parents as a cross, e.g. where the two parents
carry different favourable haplotypes at the same blocks rather than the
same one.

**When to skip it.** If you only need a parent list, not a cross list --
stop at Section 5, this adds nothing there. If you need actual contribution
numbers and a full mating plan balanced across your *whole* set of crosses
at once, with population-wide inbreeding under an explicit cap, this is not
that tool either — UC ranks each candidate cross independently with no
notion of the others, which is exactly the gap Section 9 closes next.

**In practice, choosing a `variance_model`.** Match the mode to data you
already have rather than reaching for the most rigorous one by default:
`"block_independent"` needs nothing beyond what Step 1 already produced, so
it's the right default starting point for most panels; `"phased"` is worth
the extra phasing step only if within-block gametic detail actually changes
your ranking (check by comparing the two modes' top crosses once, rather
than assuming); `"linked"` is the next step up from `"phased"` once you also
have a genetic map and suspect your target blocks are close enough together
that treating them as independent (what `"phased"` does) is a real
concern — no extra package needed, just more computation (Monte Carlo);
`"simplemating"` is worth its extra data requirements (0/2-coded DH/RIL
genotypes, a genetic map) specifically when you need the multi-generation
RIL/DH population variance itself, not just a single cross's F1-style
segregation variance. The block-independent mode below reuses exactly what
Section 3 already computed — no new model fit needed:

```{r usefulness-criterion}
uc <- usefulness_criterion(
  parent_ids           = ga_sel$selected,
  gebv                 = pred$gebv,
  selected_proportion  = 0.1,
  variance_model        = "block_independent",
  block_importance      = top_blocks,
  local_gebv            = pred$local_gebv,
  segregation_factor    = 0.5,
  verbose               = FALSE
)

nrow(uc)                                        # one row per candidate pair
head(uc[, c("parent1", "parent2", "mid_parent_gebv",
            "predicted_variance", "UC", "rank")], 10L)
```

`selected_proportion` is the fraction of each cross's progeny you would
actually keep after making it (drives the selection-intensity term `i`); set
it to whatever your programme's realistic within-cross selection pressure is.
Passing `n_progeny` (a real, finite biparental population size, e.g. `200`)
switches `i` from the asymptotic Falconer & Mackay (1996) formula to an
exact finite-population Monte Carlo estimate — the asymptotic formula
*overstates* intensity for small crosses, so this matters for realistic
biparental population sizes.

This re-ranks the *same* 20 GA-selected parents by cross potential rather
than individual merit. Worth checking directly: does the highest-UC cross
actually pair the two highest-merit parents, or something more
complementary?

```{r uc-top-cross}
top_cross <- uc[uc$rank == 1L, ]
top_cross[, c("parent1", "parent2", "mid_parent_gebv", "predicted_variance", "UC")]
pred$gebv[c(top_cross$parent1, top_cross$parent2)]      # their individual GEBVs
range(pred$gebv[ga_sel$selected])                        # vs. the full shortlist's range
```

If the top-ranked cross's parents are not simply the two highest-GEBV
individuals in the shortlist, that is UC doing its job: predicting that a
complementary pairing will segregate a better best-progeny tail than the
mid-parent value alone would suggest.

Like Section 5, `usefulness_criterion()` still does not decide *how many*
progeny to advance from each cross or manage population-wide inbreeding --
that is Section 9's job.

---

## 9. Step 7: Turning the shortlist into an actual mating plan with inbreeding under control

This is the gap Section 5.4 flagged explicitly: none of `select_parents_ga()`,
`truncation_selection()`, or `select_parents_by_family()` assigns
differential contributions or manages population-wide relatedness, and
Section 8's UC ranking scores crosses independently, with no constraint
tying them into a coherent whole-population plan. `select_parents_ocs()`
solves the actual classical Optimal Contribution Selection problem
(Meuwissen 1997): how much should each parent contribute, and which
specific matings should be made, to maximise genetic merit subject to an
explicit cap on the resulting population's relatedness. Like
`usefulness_criterion()` in Section 8, it is a downstream consumer of a
shortlist, not tied to any one shortlist-generating function: it takes
`merit` and `G` for whichever parent set you supply (named to match), plus
an optional `family` grouping, and does not itself take Section 8's UC
ranking as an input (see Section 5.5).

This package does not implement an OCS solver itself — it wraps three
mature external engines, chosen via `engine`. Two solve the actual OCS
problem: the AlphaMate executable (full evolutionary-algorithm mate
allocation, requires a separately-installed binary via `alphamate_exe`), and
optiSel's own solver (`engine = "optisel"`, via `candes()`/`opticont()`/
`matings()`/`noffspring()`, Wellmann 2019 — requires only the `optiSel`
package, no external binary). A third, `engine = "simplemating"`, is a
genuinely different algorithm class — **not** true OCS — doing discrete
greedy cross prediction/selection via SimpleMating's `planCross()`/
`selectCrosses()` (requires the `SimpleMating` package, which pulls in
`optiSel` as its own transitive dependency without using it directly).

All three share a single `target_degree` diversity-vs-gain lever (Kinghorn's
frontier-degree concept, `[0, 90]`: 0 prioritises maximising merit (the
max-gain end of the frontier, accepting more relatedness), 90 prioritises
minimising relatedness (the max-diversity end)), but via genuinely different
mechanisms per engine: AlphaMate uses it natively inside its evolutionary
algorithm; the `"optisel"` engine interpolates a mean-kinship ceiling
between two `opticont()`-solved frontier extremes specific to your candidate
set; the `"simplemating"` engine approximates it by mapping onto
`selectCrosses()`'s `culling.pairwise.k` hard relatedness cutoff over the
candidate cross set, by quantile of the observed relatedness values. See
`?select_parents_ocs`'s "Engine differences" section for the full, real,
documented differences between all three.

**Which engine should you use?** If you have a working AlphaMate
executable, it remains the most complete option (native `n_parents_max`
support, continuous frontier degree). If not, `engine = "optisel"` is the
recommended default: it solves the actual OCS problem, like AlphaMate does,
just via a different solver, and needs only the `optiSel` R package (on
CRAN, no external binary). Reach for `engine = "simplemating"` specifically
when you want SimpleMating's own discrete, greedy cross-selection algorithm
-- e.g. if your programme already standardises on SimpleMating elsewhere, or
you want to compare its cross-prediction approach against the two true-OCS
engines directly — not merely as a stand-in for OCS.

**When to reach for `select_parents_ocs()` at all.** This is the tool for an
*ongoing* population-improvement or recurrent-selection programme, where
inbreeding accumulation across cycles is a real, actively managed risk
rather than a one-off concern — population-improvement schemes with
overlapping generations, programmes maintaining several breeding pools
long-term, or any programme where "how much should each parent contribute" is
a question with real consequences beyond the current cycle. It is also the
right tool whenever you need an actual mating list with contribution
numbers, not just a shortlist (Section 5) or a cross ranking (Section 8).

**When to skip it.** For a handful of one-off biparental crosses in an
early-generation line-development programme, Section 8's UC ranking is
usually enough, and a formal population-wide inbreeding cap is not yet the
binding constraint. If no engine's dependency is installed and installing
one isn't practical right now, fall back to Section 5.8's
`coancestry_weight`/`target_degree` arguments — a softer, already-in-package
way to manage relatedness at the founder-selection step, without needing
AlphaMate, optiSel, or SimpleMating at all. Note that this function's own
`target_degree` and `select_parents_ga()`'s `target_degree` (Section 5.8)
share the same `[0, 90]` scale and direction by design, but are computed by
genuinely different mechanisms — this function solves the real OCS
contribution problem at each frontier extreme; `select_parents_ga()`
interpolates a relatedness ceiling for a penalty inside its GA. Re-tune if
you switch between them.

**In practice.** Supply `merit` (GEBV or any index you trust) and `G` (any
relationship matrix you already trust — VanRaden, pedigree, blended H;
this function does not build one for you). Start with `engine = "auto"`
(which resolves to `"alphamate"` if a working executable is supplied, else
`"optisel"` — never silently to `"simplemating"`) and `target_degree = 30`
as a first pass, then move `target_degree` toward 90 if the resulting
`mating_plan`'s relatedness looks too high for your programme's tolerance, or
toward 0 if it looks too conservative and you suspect merit is being left
on the table. All three engines honour `max_contrib_per_parent` and
guarantee a no-repeated-matings mating plan by construction; `n_parents_max`
is approximated (post-hoc contribution truncation, with a message) under
`"optisel"` and not supported at all (ignored, with a message) under
`"simplemating"` — use `engine = "alphamate"` if you need that specific
constraint enforced natively.

```{r ocs-check}
have_ocs_optisel <- requireNamespace("optiSel", quietly = TRUE) &&
  all(c("candes", "opticont", "matings", "noffspring") %in%
        getNamespaceExports("optiSel"))
have_ocs_optisel   # this section's code only runs if TRUE
```

```{r ocs, eval=have_ocs_optisel}
ocs_res <- select_parents_ocs(
  merit          = pred$gebv[ga_sel$selected],
  G              = pred$G,
  family         = family_id[ga_sel$selected],
  engine         = "optisel",       # true OCS via optiSel's own solver;
                                     # see "Which engine should you use?"
                                     # above — use "simplemating" instead
                                     # if you specifically want SimpleMating's
                                     # discrete cross-selection algorithm
  n_crosses      = 15L,
  target_degree  = 30,
  allow_selfing  = FALSE,
  verbose        = FALSE
)

ocs_res$engine_used
ocs_res$ok                                       # constraints satisfied?
head(ocs_res$mating_plan)
head(ocs_res$contributors[order(-ocs_res$contributors$contribution), ])
```

`target_degree` is the one knob worth exploring manually — raise it toward
90 if the resulting `mating_plan`'s relatedness looks too high for your
programme's tolerance, lower it toward 0 if you're leaving too much merit on
the table. Section 10 turns that manual exploration into an actual visible
tradeoff curve instead of one value at a time.

---

## 10. Step 8: Seeing the actual gain-vs-diversity tradeoff explicitly

`select_parents_ga()`'s own `coancestry_weight`/`target_degree` (Section
5.8) and `select_parents_ocs()`'s `target_degree` (Section 9) all collapse
the merit-vs-relatedness tradeoff into a single number you have to choose.
`select_parents_pareto()` instead sweeps `coancestry_weight` across a grid,
reruns the GA at each value, and keeps only the Pareto-optimal points --
the sets where no other grid point achieved both higher merit *and* lower
relatedness — so you can see the actual empirical frontier and choose a
point on it deliberately.

**When to reach for it.** Use this when you're not confident what
`coancestry_weight`/`target_degree` (Section 5.8) or `select_parents_ocs()`'s
`target_degree` (Section 9) value is "right" for your programme and want to
see the actual tradeoff shape before committing to a single number --
especially the first time you're setting this policy for a programme, or
when you need to justify a specific gain-vs-diversity choice to a review
committee or collaborator (an empirical frontier is a stronger argument
than "we used weight = 1 because it seemed reasonable"). It is also useful
periodically even in an established programme, to check whether a
previously-chosen weight still sits in a sensible place on the current
cycle's frontier.

**When to skip it.** If your programme already has institutional experience
with a specific `coancestry_weight`/`target_degree` value that has worked
well before, re-sweeping every single cycle has a real compute cost (each
grid point is a full independent GA run) for limited added information --
this is not a routine per-cycle step. Under time pressure, a single Section
5.8 or Section 9 run with a sensible default is often good enough.

**In practice.** Widen `coancestry_weights` well beyond the vignette's
speed-constrained 3-point example — the package default of 6 points
(`c(0, 0.25, 0.5, 1, 2, 4)`) is a better real-analysis starting grid, and a
denser grid gives a more informative frontier shape at the cost of runtime.
Once you've picked a frontier point by eye, pull the actual parent list
from `pareto_res$runs[[...]]$selected`, as shown below.

```{r pareto-sweep}
pareto_res <- select_parents_pareto(
  value_matrix        = value_matrix,          # from Section 5.2
  n_founders          = 20L,
  strategy            = "no_selfing",
  block_weights       = top_blocks$var_scaled,
  G                   = pred$G,
  coancestry_weights  = c(0, 1, 4),             # narrowed for vignette build
                                                 # speed; package default
                                                 # sweeps 6 grid points
  merit               = pred$gebv,
  popSize             = 100L,
  maxiter             = 200L,
  run                 = 50L,
  n_reps              = 2L,                     # reduced for vignette build
                                                 # speed (default 3)
  seed                = 1L,
  verbose             = FALSE
)

pareto_res$frontier[, c("coancestry_weight", "mean_merit",
                        "mean_relationship", "pareto_optimal")]
```

Every row is Pareto-optimal by construction here only if no grid point was
strictly dominated by another; with a narrow 3-point grid (for vignette build
speed) most or all rows typically qualify — widen `coancestry_weights` in a
real analysis for a genuinely informative frontier shape. `pareto_res$runs`
holds the full `select_parents_ga()` output (including `$selected`) for
every grid point, so once you've picked a point off the frontier by eye, the
actual parent list is `pareto_res$runs[[pareto_res$frontier$run_index[i]]]$selected`.

---

## 11. Step 9: Sanity-checking a heuristic mating plan against the true optimum

Every mate-allocation tool used or wrapped in this package — the GA search,
AlphaMate's evolutionary algorithm, optiSel's `opticont()`/`matings()`
numerical solvers, SimpleMating's `planCross()`/`selectCrosses()` — is a
**heuristic** in the sense that matters here: none is guaranteed to find the
true best *discrete mating plan* (optiSel's continuous contribution-level
optimum is exact for that sub-problem, but converting it to an actual
discrete Sire x Dam list, and enforcing per-parent/no-repeat constraints
alongside it, is still solved numerically, not exactly).
`validate_crosses_exact()` solves the same core cross-selection problem
exactly, via binary integer linear programming (`lpSolve`), so a heuristic
plan can be checked against a genuine optimum on a small-enough candidate
set. This is a validation tool, not a production replacement — integer
programming does not scale to the size of a full candidate cross list the
way a GA or greedy heuristic does (see `max_vars`).

**When to reach for it.** Use this when you want a documented, defensible
answer to "how good is our heuristic mating plan, really?" — for internal
QA, a funder or reviewer report, or before adopting a new tool (the GA,
AlphaMate, optiSel, SimpleMating's `selectCrosses()`) as your programme's
standard. It works best when your
candidate cross list is small enough to solve exactly (tens to low hundreds
of crosses after any culling — see `max_vars`), and when you're deciding
between two heuristic plans and want a common yardstick rather than
comparing their own self-reported objectives.

**When to skip it.** This is not a routine, cycle-to-cycle production step
-- for day-to-day mating-plan generation, use Section 9's engines directly.
Don't reach for it on a large, unfiltered candidate set either; pre-filter
first (a tighter `culling_pairwise_k`, or restrict to your top-ranked
crosses by `criterion_col`) rather than raising `max_vars` to force a big
problem through.

**In practice.** Feed it the same `usefulness_criterion()` output (or any
cross table with a criterion column) you'd otherwise hand to a heuristic
tool, matching whatever `max_cross`/`n_cross` constraints your real plan
used, and compare `gap_pct`. A large gap is a signal to reconsider the
heuristic tool or its settings — not necessarily a signal to switch to ILP
as your production method, since this tool's scaling limits make that
impractical for a full programme.

```{r exact-check}
have_lpsolve <- requireNamespace("lpSolve", quietly = TRUE)
have_lpsolve   # this section's code only runs if TRUE
```

```{r exact-validate, eval=have_lpsolve}
uc_top <- utils::head(uc[!is.na(uc$UC), ], 40L)   # keep the ILP small/fast

naive_top15 <- utils::head(uc_top[order(-uc_top$UC), ], 15L)  # naive: just
                                                                # take the top
                                                                # 15 by UC

exact_res <- validate_crosses_exact(
  data           = uc_top,
  n_cross        = 15L,
  max_cross      = 3L,          # no parent in more than 3 of the 15 crosses
  criterion_col  = "UC",
  heuristic_plan = naive_top15,
  verbose        = FALSE
)

exact_res$exact_objective
exact_res$heuristic_objective
exact_res$gap_pct
```

This is exactly why the check exists: "take the top 15 crosses by UC" can
violate `max_cross` outright (the same strong parent showing up in many of
the highest-UC crosses), in which case `naive_top15` isn't even a feasible
plan under the constraint and `gap_pct` will be substantial; on a panel
where the top-UC crosses happen to already be well spread across parents,
`gap_pct` can come out near zero, which is itself a useful result — it
tells you the heuristic left little to no gain on the table for this
particular candidate set. Either way, this is the same per-parent-
contribution-concentration risk Section 6's family-balance check was
watching for at the parent-selection stage, now checked quantitatively at
the mating-plan stage.

---

## 12. Step 10: Building a diverse founder set instead of a merit-ranked one

Every tool up to this point treats diversity as a *constraint* layered on
top of a merit objective. Sometimes the actual goal is the reverse — a
genebank core collection, a broad training/reference panel, or a
diversification-focused founder set where representing the panel's genetic
diversity **is** the objective, not a side constraint. `select_core_collection()`
implements the classical core-collection problem (Schoen & Brown 1993) via
farthest-point/maximin greedy traversal (Gonzalez 1985, a proven
2-approximation), with an optional merit floor so you can still ask for "the
most diverse subset among candidates that already clear a bar":

**When to reach for it.** Genebank or germplasm curation — choosing a
representative core subset from a large accession collection for long-term
conservation or characterisation; building or refreshing a broad training/
reference population for genomic prediction, where representing the full
diversity of the panel matters more than merit; or starting a new
diversification or pre-breeding programme from a wide genetic base,
deliberately before any selection pressure is applied.

**When to skip it.** If you're choosing crossing parents for the *next*
cycle of an active improvement programme, merit matters there and this tool
optimises diversity alone — Sections 5, 8, and 9 are the merit-driven tools
for that job (the `merit`/`min_sel_value` floor here only sets a minimum
bar, it does not rank by merit above that bar). Also worth pausing on if
your relationship matrix `G` is not yet trustworthy (poor marker density,
wrong ploidy assumptions) — the maximin/mean-distance traversal is only as
good as the distances it's given.

**In practice.** Choose `"maximin"` (default) if avoiding near-duplicate
accessions is the priority; choose `"mean_distance"` if you want the
selection spread evenly across the whole diversity space rather than
anchored on the most extreme outliers. The output's `trace` shows how
diversity accumulated step by step — useful for deciding whether `n_core`
is already large enough or the marginal diversity gain per added individual
is still climbing.

```{r core-collection}
core_res <- select_core_collection(
  G             = pred$G,
  n_core        = 15L,
  type          = "relationship",
  strategy      = "maximin",
  merit         = pred$gebv,
  min_sel_value = 0.5,
  min_sel_mode  = "percentile",   # keep only the top 50% by pred$gebv first
  seed          = 1L,
  verbose       = FALSE
)

core_res$mean_distance
core_res$min_distance
sort(core_res$selected)
length(intersect(core_res$selected, ga_sel$selected))   # overlap with Section 5's block-coverage set
```

A low overlap with `ga_sel$selected` is expected and informative, not a
contradiction — Section 5's GA set is engineered for joint favourable-block
coverage among above-average performers, while this set is engineered for
spread across the whole genomic diversity space. They answer different
questions and are not meant to converge on the same list.

---

## 13. The parent-selection toolkit at a glance

| Question you're asking | Use | Section |
|---|---|---|
| "What's the simplest defensible parent shortlist?" | `truncation_selection()` | 5.1 |
| "Which *set* of parents jointly covers my target haplotype blocks?" | `select_parents_ga()` | 5.2 |
| "I want block coverage, but merit to keep pulling on the search too (noise-aware)" | `select_parents_ga(merit_weight = ...)` or `merit_priority = ...` (auto-calibrated, `suggest_merit_weight()`) | 5.8 |
| "I want a relatedness cap on the GA's founder set, without guessing a raw penalty weight" | `select_parents_ga(coancestry_weight = ...)` or `target_degree = ...` (same 0-90 dial as `select_parents_ocs()`) | 5.8 |
| "My shortlist is naturally 'best few families, best few lines each'" | `select_parents_by_family()` | 5.8 |
| "Which specific *crosses* among my shortlisted parents are worth making?" | `usefulness_criterion()` | 8 |
| "How much should each parent contribute, and who mates whom, with inbreeding under an explicit cap?" | `select_parents_ocs()` | 9 |
| "What does the actual gain-vs-diversity tradeoff curve look like, rather than one dial setting?" | `select_parents_pareto()` | 10 |
| "How close is my heuristic mating plan to the true best achievable plan?" | `validate_crosses_exact()` | 11 |
| "I want a diverse founder/reference panel, not a merit-ranked shortlist" | `select_core_collection()` | 12 |

The three shortlist-generating functions among the rows above --
`truncation_selection()`, `select_parents_ga()` (the merit- and
relatedness-dial rows are variants of this same function, not separate
tools), and `select_parents_by_family()` — are *alternatives to each
other* for the same "who are my candidate parents" question, not a
sequence. Every row from `usefulness_criterion()` onward is a downstream
consumer of whichever shortlist you settled on from those three (or any
other list you supply), not specifically the output of
`select_parents_ga()` — see Section 5.5 for the full explanation of how
these two layers relate.

As a rough guide by breeding-programme shape: population-improvement or
diversity-management programmes (many families, ongoing recurrent selection)
tend to lean on Sections 9-10 (`select_parents_ocs()`/`select_parents_pareto()`)
to manage population-wide inbreeding over cycles; programmes built around a
few elite biparental families pushed toward rapid fixation lean more on
Sections 5 and 8 (`select_parents_ga()`/`usefulness_criterion()`) to pick a
tight, complementary parent/cross set. Section 12's core-collection tool
serves a different need again — building or refreshing a diverse base
population rather than either of the above. Section 11 is a check you can
run against any of the others' output, not a competing strategy.

---

## 14. Translating this into a breeding decision

Putting Sections 3-6 together into an actual shortlist for the next crossing
cycle:

```{r final-shortlist}
final_parents <- data.frame(
  Genotype       = ga_sel$selected,
  SelectionIndex = index_df$SelectionIndex[match(ga_sel$selected, index_df$id)],
  Family         = family_id[ga_sel$selected],
  WholeGenomeGEBV = pred$gebv[ga_sel$selected],
  stringsAsFactors = FALSE
)
final_parents[order(-final_parents$SelectionIndex), ]
```

This table — not the block-importance table, not the PCA plot — is the
founder-selection deliverable of this workflow: a defensible set of parents,
with the evidence for why they were chosen (block coverage in Section 5,
diversity and family balance in Section 6, and, where phased data and
genomicSimulation are available, projected genetic gain in Section 7) rather
than a single unexplained ranked list. Sections 8-12 build on top of this
same shortlist to answer the next question down — which crosses to make
(Section 8), how to turn that into a full contribution-and-mating plan with
inbreeding under control (Section 9), how to see the gain-vs-diversity
tradeoff explicitly (Section 10), how to check a heuristic mating plan
against the true optimum (Section 11), or how to build a diversity-first
founder set instead when that is the actual goal (Section 12). Section 13's
table is the fastest way back to whichever of those you need next.

---

## 15. See also

- **The HapBlockR Breeder's Guide** (`HapBlockR_Breeder_Guide.pdf`) — a
  standalone, non-technical companion covering the same seven tools from
  Sections 5 and 8-12 (what each is, when to reach for it, when to be
  cautious, how it works in practice, plus a programme-shape decision guide
  and glossary), with no R code, for breeders, programme managers, or
  reviewers who won't run this vignette themselves. Distributed as a PDF
  (not an editable Word document) so it reaches readers as a fixed
  reference. A `.pdf` has no vignette engine, so it isn't indexed by
  `vignette()` the way this document is — open it directly from the
  package's GitHub repository, or, once the package is installed, run
  `HapBlockR::open_breeder_guide()`.
- **Introduction to HapBlockR** — LD block detection itself, and the
  haplotype-stacking / association-testing tools (`score_favorable_haplotypes()`,
  `summarize_parent_haplotypes()`, `test_block_haplotypes()`) that sit
  alongside this parent-selection workflow.
- **Statistical Phasing with Beagle 5.x in HapBlockR** — how to get the
  phased `hap1`/`hap2` haplotypes Section 7's forward simulation (and
  Section 8's `"phased"` UC variance mode) requires.
- **LD Metrics: Standard r² and Kinship-Adjusted rV²** — choosing between
  the two LD metrics for family-structured or related breeding populations,
  relevant to how the blocks used throughout this vignette are detected in
  the first place.
- **HapBlockR Workflow: From Genotype File to Haplotype Features** and
  **Large-Scale Analysis: GDS and PLINK BED Backends** — running this same
  workflow end-to-end from raw genotype files, and at genome-wide marker
  density.
- `?usefulness_criterion`, `?select_parents_ocs`, `?select_parents_pareto`,
  `?validate_crosses_exact`, `?select_core_collection` — full argument
  documentation, references, and (for `select_parents_ocs()`) the honest
  verification-confidence notes on each external engine, for the five
  functions introduced in Sections 8-12.
