---
title: "From Genotypes to a Validated Breeding Decision"
author: "HapBlockR Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{From Genotypes to a Validated Breeding Decision}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(HapBlockR)
set.seed(42)
```

# Purpose

This executable vignette is a small specification of the primary workflow.
It uses included deterministic data, prints current result schemas, and
validates the final decision object. File-backed, external-tool, and
large-data routes are covered in their own integration workflows.

# Load and inspect the data

```{r data}
data(ldx_geno)
data(ldx_snp_info)
data(ldx_blocks)
data(ldx_blues)

dim(ldx_geno)
head(ldx_snp_info)
head(ldx_blocks)
```

The genotype matrix contains individuals in rows and variants in columns.
Row and column identifiers are part of the scientific identity and must be
unique and stable.

```{r identity}
stopifnot(
  !is.null(rownames(ldx_geno)),
  !is.null(colnames(ldx_geno)),
  !anyDuplicated(rownames(ldx_geno)),
  !anyDuplicated(colnames(ldx_geno)),
  all(ldx_snp_info$SNP %in% colnames(ldx_geno))
)
```

For an external file, use `read_geno()` and retain its import and cache
manifest. Memory behaviour depends on the format and selected backend.

```{r external-import, eval=FALSE}
imported <- read_geno(
  "genotypes.vcf.gz",
  multiallelic = "split",
  verbose = TRUE
)
```

# Haplotype prediction

```{r prediction}
prediction <- run_haplotype_prediction(
  geno_matrix = ldx_geno,
  snp_info = ldx_snp_info,
  blocks = ldx_blocks,
  blues = setNames(ldx_blues$YLD, ldx_blues$id),
  seed = 42,
  min_reliability = 0.30,
  verbose = FALSE
)

names(prediction)
head(prediction$block_importance)
head(prediction$gebv_uncertainty)
```

Reliability is `1 - PEV / Vg` where the fitted method provides prediction
error variance. When a model does not estimate reliability, the result records
it as unavailable so the breeder can apply the programme's evidence policy.

# Leakage-aware validation

```{r validation}
cv <- cv_haplotype_prediction(
  geno_matrix = ldx_geno,
  snp_info = ldx_snp_info,
  blocks = ldx_blocks,
  blues = setNames(ldx_blues$YLD, ldx_blues$id),
  k = 3,
  seed = 42,
  validation = "random",
  verbose = FALSE
)

cv$pa_pooled
head(cv$gebv_all)
validate(cv)
```

Random folds estimate interpolation within a reasonably exchangeable
population and are used for this demonstration. Grouped validation estimates
transfer across families, populations or sites, while forward validation
estimates prediction into later cycles, years or environments.

```{r grouped-validation, eval=FALSE}
cv_grouped <- cv_haplotype_prediction(
  geno_matrix, snp_info, blocks, blues,
  k = 5,
  validation = "grouped",
  groups = family_id,
  seed = 42
)
```

# Parent selection

```{r selection}
top_blocks <- select_top_blocks(
  prediction$block_importance,
  n = min(10L, nrow(prediction$block_importance))
)

value_matrix <- prediction$local_gebv[
  , top_blocks$block_id, drop = FALSE
]

selection <- select_parents_ga(
  value_matrix = value_matrix,
  n_founders = 8,
  seed = 42,
  n_reps = 3,
  verbose = FALSE
)

selection$selected
selection$stability
validate(selection)
```

The validation status, exact founder count, warnings, exclusions, seed,
input hashes, and stability evidence should be retained with the decision.

# Result contract

```{r contract}
names(selection$result_contract)
summary(selection)
as.data.frame(selection)
```

Promote a result after `validate()` reports passing gates. Retain fallbacks,
excluded candidates and uncertainty with the breeding decision.

# Session information

```{r session}
packageVersion("HapBlockR")
sessionInfo()
```
