---
title: "Programme-Level Models, Feasibility, and Data Exchange"
author: "HapBlockR Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Programme-Level Models, Feasibility, and Data Exchange}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(HapBlockR)
set.seed(42)
```

# Externally analysed targets

HapBlockR does not analyse raw plot records or fit replicate, block, row and
column effects. Supply genotype-level estimates from the approved external
analysis, with uncertainty. Units may be retained but are optional.

```{r targets}
external_values <- data.frame(
  id = paste0("G", 1:4),
  trait = "yield",
  value = c(5.2, 4.8, 5.0, 4.6),
  SE = c(0.20, 0.30, 0.25, 0.35)
)

targets <- prepare_breeding_targets(
  external_values,
  input_type = "BLUE",
  se_col = "SE"
)
targets$targets[
  , c("id", "value", "model_value", "precision_weight")
]
```

Accepted types are `adjusted_mean`, `BLUE`, `BLUP_identity`, `PBLUP`, `BV`,
`GCA` and `TGV`; `breeding_target_types()` gives their complete definitions.
Random predictions are deregressed from reliability or prediction error
variance. Pedigree BLUPs require the named numerator relationship matrix
`A`. External genomic predictions and external selection-index values are
rejected because HapBlockR estimates genomic effects and the final index
internally.

# Multi-trait index

Improvement directions are model inputs. Trait units are optional metadata.

```{r index}
values <- cbind(
  yield = c(5.2, 4.8, 5.0, 4.6),
  disease = c(2.0, 1.0, 2.5, 0.8)
)
rownames(values) <- paste0("G", 1:4)
G_trait <- matrix(
  c(1.0, -0.2, -0.2, 0.7), 2, 2,
  dimnames = list(colnames(values), colnames(values))
)
P_trait <- matrix(
  c(1.5, -0.1, -0.1, 1.1), 2, 2,
  dimnames = list(colnames(values), colnames(values))
)

index <- build_selection_index(
  values,
  genetic_cov = G_trait,
  phenotypic_cov = P_trait,
  economic_weights = c(yield = 1, disease = 0.5),
  directions = c(yield = "increase", disease = "decrease"),
  units = NULL,
  method = "smith_hazel"
)
index$scores
index$coefficients
validate(index)
```

`build_selection_index()` also provides deterministic Pesek-Baker,
Desired-Gain Selection Index (DGSI) and Quadratic Genomic Selection Index
(QGSI) methods. DGSI and QGSI call DesiredGainR. DGSI runs independent
optimisation replicates and automatically returns the one with the smallest
desired-response objective; breeders review the compact stability tables
rather than choosing a run manually. QGSI requires an explicit symmetric
quadratic-weight matrix and returns candidate-specific linear, squared and
cross-product contributions.

Use `fit_multitrait_gblup()` when genetic and residual covariance must be
estimated jointly. Use `fit_gxe_gblup()` with an environment kernel for
reaction-norm predictions. Environment-specific predictions support
environment-specific parent selection. The across-environment output is the
internally estimated genomic main effect; HapBlockR does not combine
environment predictions with breeder-specified environment weights.

# Operational screening

Parent 1 is the female and parent 2 the male in directed-cross screening.

```{r screening}
status <- data.frame(
  id = c("G1", "G2", "G3", "G4"),
  female_allowed = c(TRUE, TRUE, TRUE, FALSE),
  male_allowed = c(TRUE, TRUE, FALSE, TRUE),
  fertile = TRUE,
  flowering_start = c(5, 6, 7, 7),
  flowering_end = c(7, 8, 9, 9),
  heterotic_group = c("H1", "H2", "H1", "H2"),
  subpopulation = c("S1", "S1", "S2", "S2"),
  female_capacity = c(30, 20, 20, 0),
  male_capacity = c(20, 30, 0, 20),
  total_capacity = c(30, 30, 20, 20),
  seed_available = c(30, 20, 20, 0),
  pollen_available = c(20, 30, 0, 20)
)

screen_candidate_crosses(
  rbind(c("G1", "G2"), c("G1", "G3"), c("G4", "G1")),
  status,
  require_different_heterotic_groups = TRUE
)
```

# Feasibility certificate

```{r certificate}
plan <- data.frame(
  female = c("G1", "G3"),
  male = c("G2", "G4"),
  family_size = c(20, 20),
  period = "P1"
)
certificate <- certify_mating_plan(
  plan,
  status,
  min_family_size = 20,
  required_pairs = data.frame(parent1 = "G1", parent2 = "G2"),
  subpopulation_quotas = data.frame(
    subpopulation = c("S1", "S2"),
    min_contribution = c(40, 40)
  ),
  require_different_heterotic_groups = TRUE
)
certificate$certificate
certificate$binding_constraints
validate(certificate)
```

The certificate must pass before the plan is signed. Keep the violations and
binding constraints even when the plan is feasible.

# Standards-oriented exchange

Remote breeding databases should remain outside the modelling functions.
HapBlockR exports stable tables and mappings that a programme integration
layer can send to BrAPI or align with MIAPPE.

```{r exchange}
bundle <- build_breeding_exchange(
  index,
  breeding_program_id = "BP1",
  study_id = "STUDY1",
  trial_id = "TRIAL1",
  environment_id = "ENV1",
  trait_id = "SELECTION_INDEX",
  unit = "index unit",
  ontology_id = "CO_321:0000012"
)
bundle$recommendations
bundle$field_mapping

exchange_path <- tempfile("hapblockr-exchange-")
write_breeding_exchange(bundle, exchange_path)
verified <- read_breeding_exchange(exchange_path)
stopifnot(
  identical(
    verified$recommendations$recommendation_id,
    bundle$recommendations$recommendation_id
  )
)
```

# Sensitivity

Refit the intended threshold, leave-one-environment, or leave-one-population
scenarios explicitly, then compare their result objects:

```{r sensitivity, eval=FALSE}
stability <- assess_decision_stability(
  list(
    baseline = baseline_result,
    leave_E1_out = leave_E1_result,
    leave_E2_out = leave_E2_result
  ),
  top_n = 10,
  minimum_selection_frequency = 0.67
)
```

```{r session}
packageVersion("HapBlockR")
sessionInfo()
```
