---
title: "The Full Pipeline: Genotypes to a Certified Mating Plan"
author: "HapBlockR Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{The Full Pipeline: Genotypes to a Certified Mating Plan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(HapBlockR)
set.seed(1)
```

# Purpose

Every other vignette in this package is a deep dive into one topic: LD
metrics, phasing, large-scale backends, the parent-selection toolkit,
programme-level operations. This one instead runs the whole thing, start to
finish, as one continuous pipeline on HapBlockR's own small example dataset
(`ldx_geno`, `ldx_snp_info`, `ldx_blocks`, `ldx_blues`) so the relationships
between stages -- what feeds what, what each stage's output is actually used
for next -- are visible in one place rather than left to be inferred across
seven documents.

The stage structure mirrors a real production pipeline built on this package
(a CIAT Peru rice parent-selection script, distributed with the package
source under `scripts/`), organised here into four phases: evidence,
shortlist, mate allocation, sign-off. Each stage is kept short and points to
the vignette or Breeder's Guide section that covers it in depth -- this is
the connective tissue, not a replacement for those.

```{r data}
data(ldx_geno)
data(ldx_snp_info)
data(ldx_blocks)
data(ldx_blues)

dim(ldx_geno)               # 120 individuals x 230 SNPs
table(ldx_snp_info$CHR)     # three chromosomes
head(ldx_blues)
```

# Phase 1: Evidence (stages 1-4)

## Stage 1 -- Prepare the targets

A real programme's phenotype rarely arrives as a bare numeric column: it is
the output of a field-trial analysis or genetic evaluation run outside
HapBlockR, and that analysis's estimand (an adjusted mean? a BLUP shrunk
toward zero? a breeding value?) changes how it should be used downstream.
`prepare_breeding_targets()` makes that estimand explicit and, for shrunk
random-effect predictions, deregresses them before they re-enter marker or
haplotype models. See "Externally analysed targets" in the *programme
operations* vignette and `breeding_target_types()` for the complete list of
accepted input types.

The toy `ldx_blues` already stands in for a completed external analysis --
one row per individual, a single trait (`YLD`) -- so here it is used
directly as a named vector, the simplest of the four phenotype formats
`run_haplotype_prediction()` accepts:

```{r targets}
blues_vec <- setNames(ldx_blues$YLD, ldx_blues$id)
length(blues_vec)
```

## Stage 2 -- Genotypes and LD blocks

`read_geno()` is the entry point for real files (VCF, HapMap, GDS, PLINK
BED, dosage tables), with an explicit multiallelic policy and immutable
sample/variant identity preserved throughout. `ldx_geno` is already an
in-memory dosage matrix, so this vignette skips straight to LD block
detection; `Big_LD()` (single chromosome) and `run_Big_LD_all_chr()`
(genome-wide) are the tools that would normally produce `ldx_blocks` from
`ldx_geno`/`ldx_snp_info`:

```{r blocks}
head(ldx_blocks)
identical(sort(unique(ldx_blocks$CHR)), sort(unique(ldx_snp_info$CHR)))
```

See the *intro* vignette for genome-wide detection and block summarisation,
and the *large-scale* vignette for file-backed and whole-genome-sequencing
routes.

## Stage 3 -- Fit the prediction model

`run_haplotype_prediction()` extracts haplotype alleles per block, builds
the shared haplotype-block GRM, fits marker effects (GBLUP by default), and
backsolves each individual's per-block local GEBV. The return value carries
both the whole-genome GEBV every shortlist tool below scores candidates on,
and `gebv_uncertainty` -- prediction error variance, reliability, and the
`min_reliability` recommendation gate -- so "how good is this GEBV, really"
has an answer before any decision is built on top of it.

```{r prediction}
pred <- run_haplotype_prediction(
  geno_matrix = ldx_geno,
  snp_info    = ldx_snp_info,
  blocks      = ldx_blocks,
  blues       = blues_vec,
  min_snps    = 3L,
  seed        = 1,
  verbose     = FALSE
)

head(pred$gebv_uncertainty)
dim(pred$local_gebv)
dim(pred$G)
```

## Stage 4 -- Validate with a deployment-matched design

A predictive-ability estimate is only trustworthy if the cross-validation
design matches how the model will actually be deployed. `"random"` folds
estimate interpolation within an exchangeable population (used here for
speed); `"grouped"` folds (holding whole families or sites out together)
estimate transfer across families or locations; `"forward"` folds estimate
prediction into a not-yet-observed cycle. See "Leakage-aware validation" in
the *workflow* vignette for the other two designs.

```{r cv}
cv <- cv_haplotype_prediction(
  geno_matrix = ldx_geno, snp_info = ldx_snp_info, blocks = ldx_blocks,
  blues = blues_vec, k = 3L, validation = "random", seed = 1, verbose = FALSE
)
cv$pa_mean
```

# Phase 2: Shortlist (stages 5-8)

## Stage 5 -- Select the blocks that matter

`select_top_blocks()` reduces `pred$block_importance` to the smallest
top-ranked set that explains most of the variance in local GEBV -- here, the
smallest set explaining 90%. The resulting columns of `pred$local_gebv`
become `value_matrix`, the shared input to every founder-selection tool
below.

```{r top-blocks}
top_blocks <- select_top_blocks(pred$block_importance, perc_of_total_var = 0.90)
nrow(top_blocks)

value_matrix <- pred$local_gebv[, top_blocks$block_id, drop = FALSE]
dim(value_matrix)
```

## Stage 6 -- Truncation baseline

`truncation_selection()` on `pred$gebv` is the reference point every
alternative shortlist tool below should be judged against.

```{r truncation}
n_founders <- 15L
ts_sel <- truncation_selection(pred$gebv, n_founders = n_founders)
ts_sel$selected
```

## Stage 7 -- An alternative shortlist tool

Four alternatives are available depending on which constraint actually
matters for the programme -- pure block coverage, coverage plus merit,
guaranteed family or genetic-cluster representation, or diversity itself as
the objective. This vignette runs the GA (pure block-coverage) and
family-quota tools; the *breeding decisions* vignette compares all four
against each other, including a forward-simulation check of realised genetic
gain, in depth.

```{r ga}
ga_sel <- select_parents_ga(
  value_matrix  = value_matrix,
  n_founders    = n_founders,
  strategy      = "no_selfing",
  block_weights = top_blocks$var_scaled,
  popSize = 30L, maxiter = 20L, run = 10L,
  seed = 1, n_reps = 3L, verbose = FALSE
)
ga_sel$selected
ga_sel$converged
```

```{r family-quota}
# ldx_geno/ldx_blues carry no real pedigree; a synthetic family label (12
# families of 10) stands in here so family_select_mode has something to
# group on. See "Family or genetic-cluster quota selection" in the Breeder's
# Guide for all four family_select_mode variants and the shrinkage-corrected
# family ranking behind this tool.
ids    <- rownames(ldx_geno)
family <- setNames(paste0("Fam", ceiling(seq_along(ids) / 10)), ids)

fam_sel <- select_parents_by_family(
  score = pred$gebv, family = family, n_families = 5L, n_per_family = 3L,
  family_select_mode = "count", variance_method = "anova",
  use_family_relationship = TRUE, G = pred$G, verbose = FALSE
)
fam_sel$selected
fam_sel$family_ranking[, c("family", "rank_score", "rank")]
```

## Stage 8 -- See the realised gain-vs-diversity tradeoff

`select_parents_pareto()` sweeps `select_parents_ga()`'s `coancestry_weight`
across a grid and Pareto-filters the resulting (merit, relatedness) points
into an empirical frontier, so `target_degree` in stage 11 can be set from
an observed tradeoff rather than a guess.

```{r pareto}
pareto_res <- select_parents_pareto(
  value_matrix = value_matrix, n_founders = n_founders,
  strategy = "no_selfing", block_weights = top_blocks$var_scaled,
  G = pred$G, coancestry_weights = c(0, 1, 2), merit = pred$gebv,
  popSize = 30L, maxiter = 20L, run = 10L, n_reps = 2L,
  seed = 1, verbose = FALSE
)
pareto_res$frontier[, c("coancestry_weight", "mean_merit",
                        "mean_relationship", "pareto_optimal")]
```

# Phase 3: Mate allocation (stages 9-13)

## Stage 9 -- Screen candidate crosses for operational feasibility

A cross that cannot physically be made should never reach a merit-based
ranking. `screen_candidate_crosses()` checks role, fertility, flowering
overlap, heterotic group, quarantine compatibility, forbidden pairs and
reciprocal effects up front. `ldx_geno` carries no real breeding metadata, so
a minimal illustrative status table stands in here for the GA shortlist;
see "Operational screening" in the *programme operations* vignette for the
complete set of checks.

```{r screening}
ga_status <- data.frame(
  id = ga_sel$selected,
  female_allowed = TRUE,
  male_allowed = TRUE,
  fertile = TRUE,
  flowering_start = 5,
  flowering_end = 9
)
candidate_pairs <- t(utils::combn(ga_sel$selected, 2L))
screened <- screen_candidate_crosses(candidate_pairs, ga_status)
table(screened$feasible)

feasible_pairs <- screened[screened$feasible, c("female", "male")]
```

## Stage 10 -- Rank feasible crosses by usefulness criterion

`usefulness_criterion()` scores each feasible pair -- only the pairs stage 9
actually cleared, passed in via `cross_pairs` rather than regenerated from
scratch -- by predicted mid-parent value plus predicted segregation variance
at the target blocks. `variance_model = "block_independent"` is used here
since only unphased local GEBV is available; see "Variance-model variants"
in the Breeder's Guide for the phased, linked and SimpleMating-based
alternatives and what each requires.

```{r uc}
uc_res <- usefulness_criterion(
  cross_pairs = feasible_pairs, gebv = pred$gebv,
  selected_proportion = 0.1, seed = 1,
  variance_model = "block_independent",
  block_importance = top_blocks, local_gebv = pred$local_gebv,
  verbose = FALSE
)
head(uc_res[, c("parent1", "parent2", "mid_parent_gebv", "UC", "rank")])
```

## Stage 11 -- Solve for actual optimal contributions

`select_parents_ocs()` solves the real Optimal Contribution Selection
problem (Meuwissen 1997): how much each parent should contribute, and which
specific matings deliver that, under a relatedness ceiling set by
`target_degree` (informed by stage 8's frontier). It is run here restricted
to the stage 5-8 shortlist (`merit`/`G` subset to `ga_sel$selected`), not the
full 120-individual population, so its output stays on the same candidate
pool as stage 10's ranking and stage 13's exact-validation comparison below
-- OCS is solving "how to allocate this already-shortlisted set", not
re-shortlisting from scratch. This needs the optional `optiSel` or
`SimpleMating` packages; where neither is installed, the usefulness
-criterion ranking from stage 10 stands in as the shortlist for stages 12-13
below.

```{r ocs-check, include=FALSE}
have_ocs <- requireNamespace("optiSel", quietly = TRUE)
```

```{r ocs, eval=have_ocs}
shortlist <- ga_sel$selected
ocs_res <- select_parents_ocs(
  merit = pred$gebv[shortlist], G = pred$G[shortlist, shortlist],
  family = family[shortlist],
  engine = "optisel", n_crosses = 10L, max_contrib_per_parent = 3L,
  target_degree = 30, seed = 1, verbose = FALSE
)
ocs_res$mating_plan
ocs_res$ok
```

```{r ocs-fallback, eval=!have_ocs}
message("optiSel not installed -- using the stage 10 usefulness-criterion ",
        "ranking as the mating-plan candidate list instead.")
mating_plan <- uc_res[uc_res$rank <= 10L,
                      c("parent1", "parent2", "UC")]
names(mating_plan)[3] <- "criterion"
mating_plan
```

## Stage 12 -- Certify the plan and check its sensitivity

`certify_mating_plan()` validates the complete plan against family size,
parent capacity, required pairs, period-specific capacity and subpopulation
quotas, and returns every binding constraint alongside the certificate.
`assess_decision_stability()` (not run here -- see "Sensitivity" in the
*programme operations* vignette) then compares the certified plan against
threshold, model or population-scenario variants.

```{r certificate}
plan_pairs <- if (have_ocs) ocs_res$mating_plan else mating_plan
plan <- data.frame(
  female = plan_pairs$parent1,
  male = plan_pairs$parent2,
  family_size = 20
)
plan_status <- data.frame(
  id = unique(c(plan$female, plan$male)),
  female_allowed = TRUE, male_allowed = TRUE, fertile = TRUE,
  flowering_start = 5, flowering_end = 9
)
certificate <- certify_mating_plan(plan, plan_status, min_family_size = 20)
certificate$certificate
```

## Stage 13 -- Validate against the true optimum

`validate_crosses_exact()` solves the same cross-selection problem exactly
(binary integer linear programming) on the culled candidate list and reports
the heuristic plan's percentage gap below the true optimum -- a small-scale
sanity check on stage 11, not a routine replacement for it. This needs the
optional `lpSolve` package.

```{r exact-check, include=FALSE}
have_lpsolve <- requireNamespace("lpSolve", quietly = TRUE)
```

```{r exact, eval=have_lpsolve}
exact_res <- validate_crosses_exact(
  data = uc_res, n_cross = min(10L, nrow(uc_res)), max_cross = 3L,
  heuristic_plan = plan_pairs[, c("parent1", "parent2")], verbose = FALSE
)
exact_res$exact_objective
exact_res$gap_pct
```

# Phase 4: Sign-off (stages 14-16)

## Stage 14 -- Cross-check with a diversity-first alternative

`select_core_collection()` on the same relationship matrix, and
`cluster_selection_groups()` across every strategy computed above, show
whether the merit-driven selections concentrate on one or two genetic
clusters or actually spread across the population's real structure.

```{r core-collection}
core_res <- select_core_collection(
  G = pred$G, n_core = n_founders, type = "relationship",
  strategy = "maximin", merit = pred$gebv, seed = 1, verbose = FALSE
)

selection_groups <- list(
  Truncation  = ts_sel$selected,
  GA          = ga_sel$selected,
  FamilyQuota = fam_sel$selected,
  Core        = core_res$selected
)
clust_res <- cluster_selection_groups(
  G = pred$G, groups = selection_groups,
  variance_threshold = 0.95, method = "hierarchical",
  n_clusters = 3L, verbose = FALSE
)
clust_res$table
```

## Stage 15 -- Validate the result contract

Every breeder-facing result from stage 3 onward carries a `hapblockr_result`
contract; `validate()` checks the schema, immutable identifiers, input
hashes, quality gates and decision table before a result is treated as a
recommendation. See "Interpreting quality control and uncertainty" and
"Decision sign-off" in the Breeder's Guide.

```{r validate}
validate(ga_sel)
validate(fam_sel)
validate(core_res)
validate(certificate)
validate(exact_res)
```

## Stage 16 -- Export for exchange

`build_breeding_exchange()` converts a validated result into BrAPI/MIAPPE
-oriented tables, and `write_breeding_exchange()` writes them as a
checksummed CSV bundle a programme's own integration layer can consume. Any
`hapblockr_result` qualifies; the diversity-first core-collection result
from stage 14 is exported here because it best illustrates the
cross-clustering check from that stage. See "Standards-oriented exchange"
in the *programme operations* vignette.

```{r exchange}
bundle <- build_breeding_exchange(
  core_res,
  breeding_program_id = "DEMO", study_id = "DEMO-CYCLE-1", trial_id = "DEMO-T1"
)
head(bundle$recommendations[, c("entity_id", "rank", "score", "recommendable")])

exchange_path <- tempfile("hapblockr-full-pipeline-")
write_breeding_exchange(bundle, exchange_path)
list.files(exchange_path)
```

# Where each stage is covered in depth

| Stage | Vignette / Guide section |
| :--- | :--- |
| 1 | *Programme operations* vignette, "Externally analysed targets" |
| 2 | *Intro* vignette; *Large-scale* vignette for file-backed/WGS routes |
| 3-4 | *Workflow* vignette |
| 5-8 | *Breeding decisions* vignette (all four shortlist tools compared) |
| 9, 12 | *Programme operations* vignette, "Operational screening"/"Feasibility certificate" |
| 10-11, 13 | *Breeding decisions* vignette; Breeder's Guide "The nine decision tools and their variants" |
| 14 | *Breeding decisions* vignette |
| 15 | Breeder's Guide "Interpreting quality control and uncertainty" |
| 16 | *Programme operations* vignette, "Standards-oriented exchange" |

```{r session}
packageVersion("HapBlockR")
sessionInfo()
```
