---
title: "Programme-Level Models, Feasibility, and Data Exchange"
author: "HapBlockR Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Programme-Level Models, Feasibility, and Data Exchange}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(HapBlockR)
set.seed(42)
```

# Externally analysed targets

HapBlockR does not analyse raw plot records or fit replicate, block, row and
column effects. Supply genotype-level estimates from the approved external
analysis, with uncertainty. Units may be retained but are optional.

```{r targets}
external_values <- data.frame(
  id = paste0("G", 1:4),
  trait = "yield",
  value = c(5.2, 4.8, 5.0, 4.6),
  SE = c(0.20, 0.30, 0.25, 0.35)
)

targets <- prepare_breeding_targets(
  external_values,
  input_type = "BLUE",
  se_col = "SE"
)
targets$targets[
  , c("id", "value", "model_value", "precision_weight")
]
```

Accepted types are `adjusted_mean`, `BLUE`, `BLUP_identity`, `PBLUP`, `BV`,
`GCA` and `TGV`; `breeding_target_types()` gives their complete definitions.
Random predictions are deregressed from reliability or prediction error
variance. Pedigree BLUPs require the named numerator relationship matrix
`A`. External genomic predictions and external selection-index values are
rejected because HapBlockR estimates genomic effects and the final index
internally.

# Multi-trait index

Improvement directions are model inputs. Trait units are optional metadata.

```{r index}
values <- cbind(
  yield = c(5.2, 4.8, 5.0, 4.6),
  disease = c(2.0, 1.0, 2.5, 0.8)
)
rownames(values) <- paste0("G", 1:4)
G_trait <- matrix(
  c(1.0, -0.2, -0.2, 0.7), 2, 2,
  dimnames = list(colnames(values), colnames(values))
)
P_trait <- matrix(
  c(1.5, -0.1, -0.1, 1.1), 2, 2,
  dimnames = list(colnames(values), colnames(values))
)

index <- build_selection_index(
  values,
  genetic_cov = G_trait,
  phenotypic_cov = P_trait,
  economic_weights = c(yield = 1, disease = 0.5),
  directions = c(yield = "increase", disease = "decrease"),
  units = NULL,
  method = "smith_hazel"
)
index$scores
index$coefficients
validate(index)
```

`build_selection_index()` also provides deterministic Pesek-Baker,
Desired-Gain Selection Index (DGSI) and Quadratic Genomic Selection Index
(QGSI) methods. DGSI and QGSI call DesiredGainR. DGSI runs independent
optimisation replicates and automatically returns the replicate chosen by its
declared holdout or independent-validation rule; breeders review the compact
stability tables rather than choosing a run manually. QGSI requires an explicit
symmetric quadratic-weight matrix and returns candidate-specific linear,
squared and cross-product contributions.

For every method, `directions` defines favourable orientation.
`economic_weights` and `desired_gains` are non-negative magnitudes in that
oriented space; negative objectives are rejected. DGSI reports both its
model-expected transmitted genetic response in original trait units and its
realised selected-set differential in candidate SD units. DGSI and QGSI use
the selection intensity for the number of candidates actually selected. QGSI
reports model-expected gains using the complete QGSI variance and retains its
linear and quadratic weights separately because no single global QGSI
coefficient vector exists. HapBlockR converts DesiredGainR's reported response
from its analysis scale back to the original trait units. Additional QGSI
engine controls, such as explicit `Gamma`, relationship information or trait
scaling, can be supplied through `qgsi_control`. All methods expose a common
coefficient-table schema.

DGSI interprets `desired_gains` in candidate standard deviations even when
`scale_traits = FALSE`; Pesek-Baker uses original trait units. For example,
a DGSI target of 0.5 requests a selected-set shift of half a candidate standard
deviation in the favourable direction. This target is distinct from the
model-expected response transmitted to the next generation.

Both control lists require exact DesiredGainR argument names. Supply reference
data with the original-unit trait columns and an `id` column. For QGSI, `Gamma`
also uses original trait units, whereas the linear and quadratic weights refer
to the oriented, optionally scaled analysis space. `genetic_cov` and
`phenotypic_cov` remain upstream context for QGSI; they do not replace `Gamma`.

For a DGSI result, use `coefficients_original_units` when combining original-unit
marker, haplotype or block effects. Candidate scores can be reconstructed as
`trait_values %*% coefficients_original_units + score_intercept`.
The existing `coefficients$coefficient` field retains DesiredGainR's
analysis-scale coefficients. Use `engine_result` when passing the fitted index
to DesiredGainR's comparison tools.

Use `fit_multitrait_gblup()` when genetic and residual covariance must be
estimated jointly. A supplied full sampling covariance is the complete
record-error covariance by default; request
`sampling_covariance_mode = "sampling_plus_residual"` only when an additional
residual nugget is intended. Sampling-covariance keys and diagonal precision
are checked before fitting. Use `fit_gxe_gblup()` with an environment kernel for
reaction-norm predictions. Environment-specific predictions support
environment-specific parent selection. The across-environment output is the
internally estimated genomic main effect; HapBlockR does not combine
environment predictions with breeder-specified environment weights. Both
models report REML likelihoods with residual degrees of freedom; compare
those likelihoods only between models with the same fixed-effect design.

# Operational screening

Parent 1 is the female and parent 2 the male in directed-cross screening.

```{r screening}
status <- data.frame(
  id = c("G1", "G2", "G3", "G4"),
  female_allowed = c(TRUE, TRUE, TRUE, FALSE),
  male_allowed = c(TRUE, TRUE, FALSE, TRUE),
  fertile = TRUE,
  flowering_start = c(5, 6, 7, 7),
  flowering_end = c(7, 8, 9, 9),
  heterotic_group = c("H1", "H2", "H1", "H2"),
  subpopulation = c("S1", "S1", "S2", "S2"),
  female_capacity = c(30, 20, 20, 0),
  male_capacity = c(20, 30, 0, 20),
  total_capacity = c(30, 30, 20, 20),
  seed_available = c(30, 20, 20, 0),
  pollen_available = c(20, 30, 0, 20)
)

screen_candidate_crosses(
  rbind(c("G1", "G2"), c("G1", "G3"), c("G4", "G1")),
  status,
  require_different_heterotic_groups = TRUE
)
```

# Feasibility certificate

```{r certificate}
plan <- data.frame(
  female = c("G1", "G3"),
  male = c("G2", "G4"),
  family_size = c(20, 20),
  period = "P1"
)
certificate <- certify_mating_plan(
  plan,
  status,
  min_family_size = 20,
  required_pairs = data.frame(parent1 = "G1", parent2 = "G2"),
  subpopulation_quotas = data.frame(
    subpopulation = c("S1", "S2"),
    min_contribution = c(40, 40)
  ),
  require_different_heterotic_groups = TRUE
)
certificate$certificate
certificate$binding_constraints
validate(certificate)
```

The certificate must pass before the plan is signed. Keep the violations and
binding constraints even when the plan is feasible.

# Standards-oriented exchange

Remote breeding databases should remain outside the modelling functions.
HapBlockR exports stable tables and mappings that a programme integration
layer can send to BrAPI or align with MIAPPE.

```{r exchange}
bundle <- build_breeding_exchange(
  index,
  breeding_program_id = "BP1",
  study_id = "STUDY1",
  trial_id = "TRIAL1",
  environment_id = "ENV1",
  trait_id = "SELECTION_INDEX",
  unit = "index unit",
  ontology_id = "CO_321:0000012"
)
bundle$recommendations
bundle$field_mapping

exchange_path <- tempfile("hapblockr-exchange-")
write_breeding_exchange(bundle, exchange_path)
verified <- read_breeding_exchange(exchange_path)
stopifnot(
  identical(
    verified$recommendations$recommendation_id,
    bundle$recommendations$recommendation_id
  )
)
```

# Sensitivity

Refit the intended threshold, leave-one-environment, or leave-one-population
scenarios explicitly, then compare their result objects:

```{r sensitivity, eval=FALSE}
stability <- assess_decision_stability(
  list(
    baseline = baseline_result,
    leave_E1_out = leave_E1_result,
    leave_E2_out = leave_E2_result
  ),
  top_n = 10,
  minimum_selection_frequency = 0.67
)
```

```{r session}
packageVersion("HapBlockR")
sessionInfo()
```
