---
title: "DesiredGainR: DGSI and QGSI Workflows"
author: "Félicien Akohoue"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DesiredGainR: DGSI and QGSI Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(DesiredGainR)
set.seed(42)
```

# Scope

DesiredGainR constructs multi-trait selection indices from genotype-level
trait values and covariance information produced by an upstream genetic
analysis. It provides: (i) a replicated, optimised Desired-Gain Selection
Index (DGSI), and (ii) a Quadratic Genomic Selection Index (QGSI) with
explicit linear and quadratic weights.

DGSI calibrates a linear score to breeder-specified desired responses. QGSI
scores nonlinear combinations of genomic estimated breeding values (GEBVs).
The package does not infer trait GEBVs or replace the upstream mixed model.

```{r data}
traits <- c("yield", "protein", "disease")
n <- 60
candidates <- data.frame(
  GenoID = paste0("G", seq_len(n)),
  Family = rep(paste0("F", 1:6), each = 10),
  yield = rnorm(n),
  protein = rnorm(n),
  disease = rnorm(n)
)
G <- cov(candidates[traits])
dimnames(G) <- list(traits, traits)
```

# DGSI

The genetic covariance matrix `G` is mandatory. Candidate covariance is never
silently described as genetic covariance. The phenotypic or index-variable
matrix `P` may be supplied directly. When omitted, an empirical working
covariance is estimated and labelled accordingly in
`covariance_provenance`.

```{r dgsi}
dgsi <- run_dgsi(
  init_data = candidates[c("GenoID", "Family")],
  cand_data = candidates,
  trait_cols = traits,
  dg = c(yield = 0.6, protein = 0.3, disease = 0.4),
  G = G,
  lower_is_better = "disease",
  n_select = 10,
  n_iter = 100,
  n_rep = 5,
  seed = 42
)

dgsi$best_replicate
dgsi$replicate_diagnostics
head(dgsi$ranked_geno)
```

All replicates are run internally. DesiredGainR selects the replicate with the
smallest objective automatically. The breeder reviews summary diagnostics,
including coefficient variation, score-rank correlation and selected-set
agreement; no manual run selection is required.

## Thresholds as eligibility

Threshold mode first determines which candidates satisfy every favourable-
direction threshold. It then ranks the eligible pool by the optimised index
and retains at most `n_select`.

```{r thresholds}
threshold_result <- run_dgsi(
  init_data = candidates[c("GenoID", "Family")],
  cand_data = candidates,
  trait_cols = traits,
  dg = c(yield = 0.6, protein = 0.3, disease = 0.4),
  G = G,
  lower_is_better = "disease",
  select_mode = "eligible_top_n",
  trait_min = c(yield = -0.8, protein = -0.8, disease = -0.8),
  n_select = 10,
  n_iter = 100,
  n_rep = 5,
  seed = 42
)
threshold_result$eligibility
```

Missing values stop the analysis by default. Complete-case filtering or mean
imputation must be requested explicitly and is retained in `missing_data`.
Independent validation values can be passed through `validation_data`; they
are evaluated after optimisation and do not leak into replicate choice.

# QGSI

QGSI uses `w' gamma + gamma' W gamma`. The breeder supplies the linear vector
`w` and a symmetric squared/cross-product matrix `W` in the transformed trait
space. DesiredGainR does not construct `W` from desired gains because squared
terms can otherwise reward large deviations in the wrong biological
direction.

```{r qgsi}
W <- matrix(
  c(
    0.10, 0.02, -0.01,
    0.02, 0.05,  0.00,
   -0.01, 0.00,  0.08
  ),
  3,
  dimnames = list(traits, traits)
)

qgsi <- run_qgsi(
  init_data = candidates[c("GenoID", "Family")],
  gebv_data = candidates,
  trait_cols = traits,
  linear_weights = c(yield = 1, protein = 0.5, disease = 0.7),
  W = W,
  lower_is_better = "disease"
)

head(qgsi$ranked_geno)
head(qgsi$linear_contributions)
head(qgsi$quadratic_contributions)
```

The contribution tables preserve each candidate's linear, squared and
cross-product terms. Hence, the nonlinear score can be interpreted without
claiming that one global additive marker-effect vector explains every
candidate.

# Reproducible reporting

Report the upstream trait model, covariance provenance, directions,
transformations, desired gains or economic weights, selection size, random
seed, optimisation diagnostics and selected-set agreement. For QGSI, also
report the complete `W` matrix and the scientific or economic rationale for
its diagonal and cross-product entries.

```{r session}
packageVersion("DesiredGainR")
sessionInfo()
```
